---
title: "TEs on NC male phased assemblies"
author: "Zoë"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    code_folding: hide
    default_style: dark
    toc_depth: 6
    fig_caption: yes
  html_document:
    toc_depth: '6'
    df_print: paged
---

```{r setup, message=FALSE}
#include=FALSE, echo=FALSE, results=FALSE, message=FALSE, warning=FALSE
# Set the root directory
knitr::opts_knit$set(root.dir = "~/OneDrive/GradSchool/3_Analysis/2022NewRefs")

# Import libraries
library("tidyverse") #so much love
library("rtracklayer")

# Colouring & themes
pubTheme <-
  theme(title = element_text(size=10), #usually 14
        text = element_text(size=8),
        plot.background = element_rect(fill="lightgrey"),
        #plot.title = element_text(hjust = 0.5), #not sure yet if I want this
        strip.background = element_rect(linetype=0,linewidth=8,
                                        fill="grey"), #facet boxes
        strip.text = element_text(size=10), #usually 11
        legend.background = element_rect(fill="grey"),
        legend.text = element_text(size=8),
        legend.title = element_text(size=10),
        panel.background = element_rect(fill="#FFFFFF"), #plot body
        panel.grid = element_blank() #plot body pt2
  )
pubColours <- c("#2d9da6","#99b700","#8a53b6","#bf4824","#152944") #for white paper
thesisColours <- c("#d84638","#65bc5c","#01a5bc","#516fc8","#373f51") #for darker background
```


# Import {.tabset}

## Summary 

Three *four* types of annotation files are used: all TEs, intact TEs only, and genes *and full MAKER annotation results*.

## Variables

```{r importVars}
# GFF Column IDs (+ types, to prevent issues with the score column)
colIDs = c("scaff", "source", "seqOnt", "start", "end", "score", "strand",
         "phase", "attr")
colTypes <- cols(col_character(), col_character(), col_character(),
                 col_integer(), col_integer(), col_character(), 
                 col_character(), col_character(), col_character())

# Window size
winsize=1000000 #1Mb by default
```

Also, scaffold lengths

```{r scaffLengths}
scaffLens <- tibble(
  Scaffold = c("A1", "A2", "A4", "XY", "NeoY"),
  NCfemLen = c(469520921, 342707286, 180207118, 
               484402057, NA),
  H1Len = c(466657150, 320470792, 168693024, 483481970, NA),
  H2Len = c(467141259, 327917628, 174131248, 343398141, 347874410)
)
```


## Data import

Separated by karyotype

```{r import NCfem}
#### Files 
NCfemFile <- "data/NeoXfem.fasta.mod.EDTA.TEanno.gff3"
femIntactFile <- "data/NeoXfem.fasta.mod.EDTA.intact.gff3"
NCfemGeneFile <- "data/NCfem.genes.gff"



####   Scaffold names   ###
NCfemScaffs <- c("Scaffold_1__2_c", "Scaffold_2__1_c", "Scaffold_4__2_c",
                 "Scaffold_3__1_c")
NCfemGeneScaffs <- c("Scaffold_1__2_contigs__length_484402057",
                   "Scaffold_2__1_contigs__length_469520921",
                   "Scaffold_4__2_contigs__length_342707286",
                   "Scaffold_3__1_contigs__length_180207118")


####   Importing.  ###
# All TEs annotated by EDTA
NCfemEDTAraw <- read_table(NCfemFile, col_names = colIDs, skip = 6,
                           col_types = cols(source=col_skip()))
# Only the intact TEs annotated by EDTA
NCfemIntactRaw <- 
  read_table(femIntactFile,col_names=colIDs, skip = 6,
             col_types = cols(source=col_skip()))
# The 'gene' entries from the MAKER annotation
NCfemGenesRaw <- read_tsv(NCfemGeneFile, col_names = colIDs,
                          col_types = cols(source = col_skip()))

#### Clean up files
rm(NCfemFile, femIntactFile, NCfemGeneFile)
```

```{r importH1}
# Files 
H1NCmaleFile <- "data/H1.fa.mod.EDTA.TEanno.gff3"
H1intactFile <- "data/H1.fa.mod.EDTA.intact.gff3"
H1geneFile <- "data/H1NCmale.genes.gff"
H1makerFile <- "data/H1NCmale.maker.gff"

# Hap1 named scaffolds
H1maleScaffs<-c("X","A1","A2","A4") # Used to be X_NeoX


# All TEs annotated by EDTA
H1EDTAraw <- read_table(H1NCmaleFile, col_names = colIDs, skip = 6,
                        col_types = colTypes)
# Only the intact TEs annotated by EDTA
H1IntactRaw <- 
  read_table(H1intactFile, col_names=colIDs, skip = 6,
             col_types = cols(source=col_skip()))
# The 'gene' entries from the MAKER annotation
H1GenesRaw <- read_tsv(H1geneFile, col_names = colIDs,
                      col_types = cols(source = col_skip()))
# All output from MAKER
H1MakerRaw <- readGFF(H1makerFile) %>%
  as_tibble() %>% #haven't figured out what additional utility might be in this rtracklayer table, so just changed it to a tibble
  rename(scaff = seqid) %>% #just to get the var name to match what I use
  filter(scaff %in% H1maleScaffs) %>% #just to make it a reasonable size
  select(!c(source, score))
#view(makertest)

# Clean up files
rm(H1NCmaleFile, H1intactFile, H1geneFile, H1makerFile)
```

```{r importH2}
# Files
H2NCmaleFile <- "data/H2.fa.mod.EDTA.TEanno.gff3"
H2intactFile <- "data/H2.fa.mod.EDTA.intact.gff3"
H2geneFile <- "data/H2NCmale.genes.gff"
H2makerFile <- "data/H2NCmale.maker.gff"


# Hap2 named scaffolds
H2maleScaffs <- c("A1", "A2", "A4", "Y1", "Y2")

# All TEs annotated by EDTA
H2EDTAraw <- read_table(H2NCmaleFile, col_names = colIDs, skip = 6,
                        col_types = colTypes)
# Only the intact TEs annotated by EDTA
H2IntactRaw <- 
  read_table(H2intactFile, col_names=colIDs, skip = 6,
             col_types = cols(source=col_skip()))
# The 'gene' entries from the MAKER annotation
H2GenesRaw <-  read_tsv(H2geneFile, col_names = colIDs,
                        col_types = cols(source = col_skip()))
# All output from MAKER
H2MakerRaw <- readGFF(H2makerFile) %>%
  as_tibble() %>% #haven't figured out what additional utility might be in this rtracklayer table, so just changed it to a tibble
  rename(scaff = seqid) %>% #just to get the var name to match what I use
  filter(scaff %in% H2maleScaffs) %>% #just to make it a reasonable size
  select(!c(source, score))


# Clean up files
rm(H2NCmaleFile, H2intactFile, H2geneFile, H2makerFile)
# Also clean up any other import things
rm(colTypes, colIDs)
```


# Clean and Summarise

## Functions

```{r cleanFunctions}
filterTEs <- function(df) {
  df %>%
    filter(seqOnt != "target_site_duplication", 
         seqOnt != "long_terminal_repeat",
         seqOnt != "repeat_region") %>%
    group_by(scaff)
}

filterGenes <- function(df) {
  df %>%
    filter(!(str_detect(attr, "[Tt]ranspos")),
           !(str_detect(attr, "[Rr]ibonuclease H")),
           !(str_detect(attr, "[Rr]etrovirus")), 
           !(str_detect(attr, "[Uu]nknown"))) %>%
    group_by(scaff)
}

filterMAKER <- function(df) {
  keepOnly <- df %>%
    filter(type == "mRNA") %>%
    filter(!(str_detect(Note, "[Tt]ranspos")),
           !(str_detect(Note, "[Rr]ibonuclease H")),
           !(str_detect(Note, "[Rr]etrovirus"))) %>%#,
#           !(str_detect(Note, "[Uu]nknown"))) %>%
    pull(Name)
  df %>%
    filter(type == "CDS") %>%
    filter(Parent %in% keepOnly)
}


renameTEs <- function(df) {
  df %>%
    mutate( seqOnt = recode(
      seqOnt, Copia_LTR_retrotransposon="RLC",Gypsy_LTR_retrotransposon="RLG",
      LTR_retrotransposon="RLX",LINE_element="RIX",helitron="DHH",
      CACTA_TIR_transposon="DTC",hAT_TIR_transposon="DTA",Mutator_TIR_transposon="DTM",
      PIF_Harbinger_TIR_transposon="DTH",Tc1_Mariner_TIR_transposon="DTT") )
}

```

```{r windowFunctions}
window <- function(df) {
  df %>%
    mutate(winnum = start %/% as.numeric(winsize),
           length = abs(end - start))
}

summarizeGenes <- function(df) {
  df %>%
    group_by(scaff, winnum) %>%
    summarise( bp = sum(length), n = n() ) %>%
    mutate(amt = bp/winsize)
}

summarizeTEs <- function(df) {
  df %>%
    group_by(scaff, winnum, seqOnt) %>%
    summarise( bp = sum(length), n = n() ) %>%
    #account for overlapping TEs leading to >100% TE density
    mutate(amt = case_when(bp>winsize ~ 1,
                           TRUE ~ bp/winsize))
}
```


## Windowing

```{r cleanFem}
NCfemEDTA <- NCfemEDTAraw %>%
  filter(scaff %in% NCfemScaffs) %>%
  filterTEs %>%
  window %>%
  summarizeTEs %>%
  renameTEs

NCfemIntact <- NCfemIntactRaw %>%
  filter(scaff %in% NCfemScaffs) %>%
  filterTEs %>%
  window %>%
  summarizeTEs %>%
  renameTEs

NCfemGenes <- NCfemGenesRaw %>%
  filter(scaff %in% NCfemScaffs) %>%
  filterGenes %>%
  window %>%
  summarizeGenes
```

```{r cleanH1, message=FALSE}
H1EDTA <- H1EDTAraw %>%
  filter(scaff %in% H1maleScaffs) %>%
  filterTEs %>%
  window %>%
  summarizeTEs %>%
  renameTEs

H1Intact <- H1IntactRaw %>%
  filter(scaff %in% H1maleScaffs) %>%
  filterTEs %>%
  window %>%
  summarizeTEs %>%
  renameTEs

H1Genes <- H1GenesRaw %>%
  filter(scaff %in% H1maleScaffs) %>%
  filterGenes %>%
  window %>%
  summarizeGenes

H1CDS <- H1MakerRaw %>%
  filterMAKER %>%
  group_by(scaff) %>%
  window %>%
  summarizeGenes
```

```{r cleanH2, message=FALSE}
H2EDTA <- H2EDTAraw %>%
  filter(scaff %in% H2maleScaffs) %>%
  filterTEs %>%
  window %>%
  summarizeTEs %>%
  renameTEs

H2Intact <- H2IntactRaw %>%
  filter(scaff %in% H2maleScaffs) %>%
  filterTEs %>%
  window %>%
  summarizeTEs %>%
  renameTEs

H2Genes <- H2GenesRaw %>%
  filter(scaff %in% H2maleScaffs) %>%
  filterGenes %>%
  window %>%
  summarizeGenes

H2CDS <- H2MakerRaw %>%
  filterMAKER %>%
  group_by(scaff) %>%
  window %>%
  summarizeGenes
```

```{r deleteRawFiles}
rm(NCfemEDTAraw, NCfemIntactRaw, NCfemGenesRaw)
rm(H1EDTAraw, H1IntactRaw, H1GenesRaw, H1MakerRaw)
rm(H2EDTAraw, H2IntactRaw, H2GenesRaw, H2MakerRaw)
```

## Divide up sex chromosomes

Split the EDTA annotations for X and Y into PAR1/2, old, and new based on window number (approximate position). Join the H1 and H2 sets of TE bp per window into *allSummedTEs*.

```{r TEsInSexRegions}
XsummedTEs <- H1EDTA %>%
  filter(scaff == "X") %>%
  mutate(region = case_when(
    winnum < 71   ~ "PAR1",
    winnum < 261  ~ "Old",
    winnum < 367  ~ "New",
    winnum >= 367 ~ "PAR2"
  )) %>%
  group_by(seqOnt, region) %>%
  summarize(summedbp = sum(bp)) %>%
  mutate(X = case_when(
    region == "PAR1"   ~ summedbp/71000000,
    region == "Old"    ~ summedbp/(261000000-71000000),
    region == "New"    ~ summedbp/(367000000-261000000),
    region == "PAR2"   ~ summedbp/(483000000-367000000)
  ))

Y1summedTEs <- H2EDTA %>%
  filter(scaff == "Y1") %>%
  mutate(region = case_when(
    winnum < 64   ~ "New",
    winnum < 272  ~ "Old",
    winnum >= 272 ~ "PAR1"
  )) %>%
  group_by(seqOnt, region) %>%
  summarize(summedbp = sum(bp)) %>%
  mutate(Y1 = case_when(
    region == "New"   ~ summedbp/64000000,
    region == "Old"    ~ summedbp/(272000000-64000000),
    region == "PAR1"    ~ summedbp/(343000000-272000000)
  ))

Y2summedTEs <- H2EDTA %>%
  filter(scaff == "Y2") %>%
  mutate(region = case_when(
    winnum < 121  ~ "PAR2",
    winnum < 159  ~ "New",
    winnum >= 159 ~ "Old"
  )) %>%
  group_by(seqOnt, region) %>%
  summarize(summedbp = sum(bp)) %>%
  mutate(Y2 = case_when(
    region == "PAR2"   ~ summedbp/121000000,
    region == "New"    ~ summedbp/(159000000-121000000),
    region == "Old"    ~ summedbp/(348000000-159000000)
  ))

allSummedTEs <- full_join(XsummedTEs, Y1summedTEs, 
                          by = join_by(seqOnt, region)) %>%
  full_join(Y2summedTEs, by = join_by(seqOnt, region)) %>%
  pivot_longer(cols = c(X, Y1, Y2), 
               names_to = "Chromosome", values_to = "bp")
```


# Visualise

## Functions

```{r plottingFunctions}
plotWindowedTEs <- function(df, name) {
  ggplot(df, aes(winnum, n, colour=seqOnt), size = 0.3) +
    geom_line() +
    pubTheme +
    labs(x = paste(name, "position (1Mb windows)"),
         y = "Number of TEs", colour = "TE Type") +
    facet_grid(vars(scaff))
}

plotGenes <- function(df, name) {
  ggplot(df, aes(winnum,n)) +
  geom_line() +
  pubTheme + 
  labs(x = paste(name, "Position (1Mb windows)"), y = "Number of genes") +
  facet_grid(vars(scaff), scale="free_x", space="free")
}

plotElements <- function(TEs, genes, name) {
  ggplot() +
  geom_line(data = TEs, aes(winnum, amt, colour=seqOnt), linewidth=0.3) +
  geom_line(data = genes, aes(winnum,amt)) +
  #ylim(0,1) +
  pubTheme +
  labs(x = paste(name, "position (1Mb windows)"), 
       y = paste(name, "Fraction of window"), colour="TE Type") +
  facet_grid(vars(scaff), scale="free_x", space="free")
}

plotFacetedTEs <- function(df, name) {
  ggplot(df, aes(winnum,n), size = 0.3) +
  geom_line() +
  pubTheme +
  labs(x = paste(name, "position (1Mb windows)"),
       y = "Number of TEs", colour = "TE Type") +
  facet_grid(vars(seqOnt), vars(scaff), space = "free", scale = "free")
}
```

```{r barPlots}
# Control number of elements for Scaffold Length
controlH1Counts <- function(df) {
  df %>%
    group_by(seqOnt, scaff) %>%
    tally() %>%
    mutate( num = case_when(
      scaff == "A1"     ~ n/scaffLens$H1Len[1],
      scaff == "A2"     ~ n/scaffLens$H1Len[2],
      scaff == "A4"     ~ n/scaffLens$H1Len[3],
      scaff == "X"      ~ n/scaffLens$H1Len[4]
    )) 
}
controlH2Counts <- function(df) {
  df %>%
    group_by(seqOnt, scaff) %>%
    tally() %>%
    mutate( num = case_when(
      scaff == "A1"   ~ n/scaffLens$H2Len[1],
      scaff == "A2"   ~ n/scaffLens$H2Len[2],
      scaff == "A4"   ~ n/scaffLens$H2Len[3],
      scaff == "Y1" ~ n/scaffLens$H2Len[4],
      scaff == "Y2" ~ n/scaffLens$H2Len[5]
    ))
}

# Divide length of elements by Scaffold Length
controlH1Len <- function(df) {
  df %>%
    group_by(seqOnt, scaff) %>%
    summarise(n = sum(bp)) %>%
    mutate( num = case_when(
      scaff == "A1"     ~ n/scaffLens$H1Len[1],
      scaff == "A2"     ~ n/scaffLens$H1Len[2],
      scaff == "A4"     ~ n/scaffLens$H1Len[3],
      scaff == "X"      ~ n/scaffLens$H1Len[4]
    )) 
}
controlH2Len <- function(df) {
  df %>%
    group_by(seqOnt, scaff) %>%
    summarise(n = sum(bp)) %>%
    mutate( num = case_when(
      scaff == "A1"   ~ n/scaffLens$H2Len[1],
      scaff == "A2"   ~ n/scaffLens$H2Len[2],
      scaff == "A4"   ~ n/scaffLens$H2Len[3],
      scaff == "Y1" ~ n/scaffLens$H2Len[4],
      scaff == "Y2" ~ n/scaffLens$H2Len[5]
    ))
}

# Make bar plot 
plotCols <- function(df, name, legPos) {
  ggplot(df, aes(x = seqOnt, y = num, fill = scaff)) +
  geom_col(position = position_dodge()) +
  pubTheme + theme(legend.position = legPos) +
  labs(x = "TE Type", y = "Number in window divided by scaffold length", 
       fill = paste(name, "scaffold"))
}

# Full chromosomal proportions
plotProportions <- function(TEs, Genes, name, legPos) {
  TEsums <- TEs %>%
    group_by(scaff) %>%
    summarize(TEs = sum(bp))
  geneSums <- Genes %>%
    group_by(scaff) %>%
    summarize(Genes = sum(bp))
  df <- left_join(TEsums, geneSums) %>%
    pivot_longer(cols = c(TEs, Genes), 
                 names_to = "element", values_to = "bp")
  
  ggplot(df, aes(x = scaff, y = bp, fill = element)) +
    geom_col() +
    pubTheme + theme(legend.position = legPos) +
    labs(x = paste(name, "chromosome"), y = "Base pairs", 
         fill = "Element type")
}
```


## TE barplots

Plot fraction of window and count data for TEs.

```{r fractionCols, message=FALSE}
H1EDTA %>%
  controlH1Len() %>%
  plotCols("H1", c(0.25, 0.7))

H2EDTA %>%
  controlH2Len() %>%
  plotCols("H2", c(0.25, 0.7))

H1Intact %>%
  controlH1Len() %>%
  plotCols("H1", c(0.25, 0.7))

H2Intact %>%
  controlH2Len() %>%
  plotCols("H2", c(0.25, 0.7))
```

```{r countCols, message=FALSE}
H1EDTA %>%
  controlH1Counts() %>%
  plotCols("H1", c(0.25, 0.7))

H2EDTA %>%
  controlH2Counts() %>%
  plotCols("H2", c(0.25, 0.7))

H1Intact %>%
  controlH1Counts() %>%
  plotCols("H1", c(0.25, 0.7))

H2Intact %>%
  controlH2Counts() %>%
  plotCols("H2", c(0.25, 0.7))
```

## Sex chromosome proportions

```{r plotSummedTEs}
ggplot(allSummedTEs, aes(x = region, y = bp, fill = Chromosome)) +
  geom_col(position = position_dodge()) +
  pubTheme + scale_fill_manual(values = pubColours) +
  labs(x = "Region", y = "Fraction of region occupied by TEs", 
       fill = "Chromosome") #+
  #facet_grid(vars(seqOnt))

ggplot(allSummedTEs, aes(x = Chromosome, y = bp, fill = region)) +
  geom_col(position = position_dodge()) +
  pubTheme + scale_fill_manual(values = pubColours) +
  labs(x = "Chromosome", y = "Fraction of region occupied by TEs", 
       fill = "Region")

allSummedTEs %>%
  #filter(seqOnt == "DTM") %>%
  ggplot(aes(x = seqOnt, y = bp, fill = Chromosome)) +
  geom_col(position = position_dodge()) +
  pubTheme + scale_fill_manual(values = pubColours) +
  #theme(legend.position = c(0.2,0.8)) +
  labs(x = "Region", y = "Fraction of region occupied by TEs", 
       fill = "Chromosome") +
  facet_grid(vars(region))
```

## Fraction of window annotated with Genes, TEs

```{r plotElementDensities}
plotElements(H1EDTA, H1Genes, "H1")
plotElements(H2EDTA, H2Genes, "H2")
```

# Testing things

## CDS instead of gene?

```{r testingGeneFiltering}
H1MakerRaw %>%
  filter(type == "CDS") %>%
  group_by(scaff) %>%
  window %>%
  summarizeGenes %>%
  ggplot() +
  geom_line(data = H1Genes, aes(winnum, amt), colour = pubColours[1], size = 1.1) +
  geom_line(aes(winnum, amt), colour = pubColours[3], size = 1.2) +
  geom_line(data = H1CDS, aes(winnum, amt), colour = pubColours[2], size = 1.1) +
  pubTheme +
  labs(x = "H1 position (1Mb windows)",
       y = "CDS frequency") +
  facet_grid(vars(scaff), scale="free_x", space="free")
```

```{r CDSvsGene, message=FALSE}
# Plot counts
H1CDS %>%
  ggplot() +
  geom_line(data = H1Genes, aes(winnum,  n), colour = pubColours[1], size = 1.1) +
  geom_line(aes(winnum, n), colour = pubColours[2], size = 1.1) +
  pubTheme +
  labs(x = "H1 position (1Mb windows)", 
       y = "Element count") +
  facet_grid(vars(scaff), scale="free_x", space="free")

H2CDS %>%
  ggplot() +
  geom_line(data = H2Genes, aes(winnum, n), colour = pubColours[1], size = 1.1) +
  geom_line(aes(winnum, n), colour = pubColours[2], size = 1.1) +
  pubTheme +
  labs(x = "H2 position (1Mb windows)", 
       y = "Element count") +
  facet_grid(vars(scaff), scale="free_x", space="free")




# Plot densities (equivalent to bp)
H1CDS %>%
  ggplot() +
  geom_line(data = H1Genes, aes(winnum, amt), colour = pubColours[1]) +
  geom_line(aes(winnum, amt), colour = pubColours[2]) +
  pubTheme +
  labs(x = "H1 position (1Mb windows)", 
       y = "Fraction of window") +
  facet_grid(vars(scaff), scale="free_x", space="free")

H2CDS %>%
  ggplot() +
  geom_line(data = H2Genes, aes(winnum, amt), colour = pubColours[1]) +
  geom_line(aes(winnum, amt), colour = pubColours[2]) +
  pubTheme +
  labs(x = "H2 position (1Mb windows)", 
       y = "Fraction of window") +
  facet_grid(vars(scaff), scale="free_x", space="free")
```

