---
title: "TEs on NC male phased assemblies"
author: "Zoë"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    code_folding: hide
    default_style: dark
    toc_depth: 6
    fig_caption: yes
  html_document:
    toc_depth: '6'
    df_print: paged
---

```{r setup, message=FALSE}
#include=FALSE, echo=FALSE, results=FALSE, message=FALSE, warning=FALSE
# Set the root directory
knitr::opts_knit$set(root.dir = "~/2022NewRefs")

# Import libraries
library("tidyverse") #so much love
#library("rtracklayer") #just for readGFF?

# Colouring & themes
pubTheme <-
  theme(title = element_text(size=10), #usually 14
        text = element_text(size=8),
        plot.background = element_rect(fill="lightgrey"),
        #plot.title = element_text(hjust = 0.5), #not sure yet if I want this
        strip.background = element_rect(linetype=0,linewidth=8,
                                        fill="grey"), #facet boxes
        strip.text = element_text(size=10), #usually 11
        legend.background = element_rect(fill="grey"),
        legend.text = element_text(size=8),
        legend.title = element_text(size=10),
        panel.background = element_rect(fill="#FFFFFF"), #plot body
        panel.grid = element_blank() #plot body pt2
  )
pubColours <- c("#2d9da6","#99b700","#8a53b6","#bf4824","#152944") #for white paper
#thesisColours <- c("#d84638","#65bc5c","#01a5bc","#516fc8","#373f51") #for darker background
```


# Data {.tabset}

## Summary 

Three *four* types of annotation files are used: all TEs, intact TEs only, and genes *and full MAKER annotation results*.

## Setting Variables

```{r importVars}
# GFF Column IDs (+ types, to prevent issues with the score column)
colIDs = c("scaff", "source", "seqOnt", "start", "end", "score", "strand",
         "phase", "attr")
colTypes <- cols(col_character(), col_character(), col_character(),
                 col_integer(), col_integer(), col_character(), 
                 col_character(), col_character(), col_character())

# Window size
winsize=1000000 #1Mb by default
```

Also, scaffold lengths

```{r scaffLengths}
scaffLens <- tibble(
  Scaffold = c("A1", "A2", "A4", "XY", "NeoY"),
  H1Len = c(466657150, 320470792, 168693024, 483481970, NA),
  H2Len = c(467141259, 327917628, 174131248, 343398141, 347874410)
)
```


## Import

```{r importH1}
# Files 
#H1NCmaleFile <- "data/H1.fa.mod.EDTA.TEanno.gff3"
H1NCmaleFile <- "data/H1.panEDTA.gff3"
H1intactFile <- "data/H1.fa.mod.EDTA.intact.gff3"
H1geneFile <- "data/H1NCmale.genes.gff"
H1liftedFile <- "data/Hap2litofftohap1.filtered.gff"

# Hap1 named scaffolds
H1scaffs <- c("X","A1","A2","A4")


# All TEs annotated by EDTA
H1EDTAraw <- read_table(H1NCmaleFile, col_names = colIDs, skip = 6,
                        col_types = colTypes)
# Only the intact TEs annotated by EDTA
H1IntactRaw <- 
  read_table(H1intactFile, col_names = colIDs, skip = 6,
             col_types = cols(source=col_skip()))
# The 'gene' entries from the MAKER annotation
H1GenesRaw <- read_tsv(H1geneFile, col_names = colIDs,
                      col_types = cols(source = col_skip()))
# All entries lifted over from Hap2's MAKER annotation
H1LiftedRaw <- read_tsv(H1liftedFile, col_names = colIDs,
                      col_types = cols(source = col_skip()))


# Clean up files
rm(H1NCmaleFile, H1intactFile, H1geneFile, H1liftedFile)
```

```{r importH2}
# Files
#H2NCmaleFile <- "data/H2.fa.mod.EDTA.TEanno.gff3"
H2NCmaleFile <- "data/H2.panEDTA.gff3"
H2intactFile <- "data/H2.fa.mod.EDTA.intact.gff3"
H2geneFile <- "data/H2NCmale.genes.gff"
H2liftedFile <- "data/Hap1litofftohap2.filtered.gff"
#H2makerFile <- "data/H2NCmale.maker.gff"


# Hap2 named scaffolds
H2scaffs <- c("A1", "A2", "A4", "Y1", "Y2")

# All TEs annotated by EDTA
H2EDTAraw <- read_table(H2NCmaleFile, col_names = colIDs, skip = 6,
                        col_types = colTypes)
# Only the intact TEs annotated by EDTA
H2IntactRaw <- 
  read_table(H2intactFile, col_names=colIDs, skip = 6,
             col_types = cols(source=col_skip()))
# The 'gene' entries from the MAKER annotation
H2GenesRaw <-  read_tsv(H2geneFile, col_names = colIDs,
                        col_types = cols(source = col_skip()))
# All entries lifted over from Hap1's MAKER annotation
H2LiftedRaw <- read_tsv(H2liftedFile, col_names = colIDs,
                      col_types = cols(source = col_skip()))
# All output from MAKER
# H2MakerRaw <- readGFF(H2makerFile) %>%
#   as_tibble() %>% #haven't figured out what additional utility might be in this rtracklayer table, so just changed it to a tibble
#   rename(scaff = seqid) %>% #just to get the var name to match what I use
#   filter(scaff %in% H2maleScaffs) %>% #just to make it a reasonable size
#   select(!c(source, score))


# Clean up files
rm(H2NCmaleFile, H2intactFile, H2geneFile, H2liftedFile)
# Also clean up any other import things
rm(colTypes, colIDs)
```


# Clean and Summarise

## Functions

Functions to clean TE data (rename TE types to match Wicker et al., )

```{r cleanFunctions}
filterTEs <- function(df) {
  df %>%
    filter(seqOnt != "target_site_duplication", 
         seqOnt != "long_terminal_repeat",
         seqOnt != "repeat_region",
         seqOnt != "low_complexity") %>%
    group_by(scaff)
}

renameTEs <- function(df) {
  df %>%
    mutate( seqOnt = recode(
      seqOnt, Copia_LTR_retrotransposon="RLC",
      Gypsy_LTR_retrotransposon="RLG", LTR_retrotransposon="RLX", 
      LINE_element="RIX", helitron="DHH", CACTA_TIR_transposon="DTC",
      hAT_TIR_transposon="DTA", Mutator_TIR_transposon="DTM", 
      PIF_Harbinger_TIR_transposon="DTH", 
      Tc1_Mariner_TIR_transposon="DTT") )
}

filterGenes <- function(df) {
  df %>%
    filter(!(str_detect(attr, "[Tt]ranspos")),
           !(str_detect(attr, "[Rr]ibonuclease H")),
           !(str_detect(attr, "[Rr]etrovirus")), 
           !(str_detect(attr, "[Uu]nknown"))) %>%
    group_by(scaff)
}
```

```{r divideSexRegionFunctions}
splitX <- function(df) {
  df %>% mutate(region = case_when(
    winnum < 71   ~ "PAR1",
    winnum < 261  ~ "Old",
    winnum < 367  ~ "New",
    winnum >= 367 ~ "PAR2"
  )) 
}

splitY1 <- function(df) {
  df %>% mutate(region = case_when(
    winnum < 64   ~ "New",
    winnum < 272  ~ "Old",
    winnum >= 272 ~ "PAR1"
  ))
}

splitY2 <- function(df) {
  df %>% mutate(region = case_when(
    winnum < 121  ~ "PAR2",
    winnum < 159  ~ "New",
    winnum >= 159 ~ "Old"
  ))
}
```


```{r makerOnly, include=FALSE}
filterMAKER <- function(df) {
  keepOnly <- df %>%
    filter(type == "mRNA") %>%
    filter(!(str_detect(Note, "[Tt]ranspos")),
           !(str_detect(Note, "[Rr]ibonuclease H")),
           !(str_detect(Note, "[Rr]etrovirus"))) %>%#,
#           !(str_detect(Note, "[Uu]nknown"))) %>%
    pull(Name)
  df %>%
    filter(type == "CDS") %>%
    filter(Parent %in% keepOnly)
}
```

```{r windowFunctions}
window <- function(df) {
  df %>%
    mutate(winnum = start %/% as.numeric(winsize),
           length = abs(end - start))
}

summarizeGenes <- function(df) {
  df %>%
    group_by(scaff, winnum) %>%
    summarise( bp = sum(length), n = n() ) %>%
    mutate(amt = bp/winsize)
}

summarizeTEs <- function(df) {
  df %>%
    group_by(scaff, winnum, seqOnt) %>%
    summarise( bp = sum(length), n = n() ) %>%
    #account for overlapping TEs leading to >100% TE density
    mutate(amt = case_when(bp>winsize ~ 1,
                           TRUE ~ bp/winsize))
}
```

## TE Family Data

### Parse family information from EDTA

Cleaning up the raw input: renaming the TE types, filtering out the excess LTR_retriever lines, and *parsing the attributes column* to grab the Name, ID, and Identity information from it. 

This is different from the other parsed EDTA data mainly because it's not windowed? I could theoretically use this 

```{r TEfamilyParsing}
H1TEs <- H1EDTAraw %>%
  filter(scaff %in% H1scaffs) %>%
  filterTEs %>%
  renameTEs %>%
  mutate(ID = str_match(attr, "ID=([^;]+)")[,2],
         family = str_match(attr, "Name=([^;]+)")[,2]) %>%
  select( !c(phase, attr, strand) )

H2TEs <- H2EDTAraw %>%
  filter(scaff %in% H2scaffs) %>%
  filterTEs %>%
  renameTEs %>%
  group_by(scaff) %>%
  mutate(ID = str_match(attr, "ID=([^;]+)")[,2],
         family = str_match(attr, "Name=([^;]+)")[,2]) %>%
  select( !c(phase, attr, strand) )
```

### Identity value tangent

Interestingly, there were issues with the identity variable because of NA values in the source file (the TE was identified structurally and therefore didn't have a homology value).

```{r testingIdentity, eval=FALSE}
H1EDTAraw %>%
  filter(scaff %in% H1scaffs) %>%
  filterTEs %>%
  renameTEs %>%
  mutate(ID = str_match(attr, "ID=([^;]+)")[,2],
         family = str_match(attr, "Name=([^;]+)")[,2],
         identity = as.numeric(str_match(attr, "[i|I]dentity=([^;]+)")[,2])) %>%
  select( !c(phase, attr, strand) ) %>%
  filter(is.na(identity))
```

585 TEs without identity values are helitrons. I don't have any uses for the identity value right now, so I'll leave that alone.


### TE family size on sex chromosomes

The largest families for both H1 and H2 are Copia elements found on A1 + Y1/Y2/X. They're different families, but it's interesting that the pattern is shared!

```{r famSummariseByScaff}
H1TEs %>%
  filter(scaff == "X") %>%
  group_by(scaff, seqOnt, family) %>%
  summarise(n = n(), bp = sum(end - start), .groups = "keep") %>%
  group_by(scaff, seqOnt) %>%
  arrange(desc(bp)) %>%
  head(50)

H2TEs %>%
  filter(scaff == "Y1" | scaff == "Y2") %>%
  group_by(scaff, seqOnt, family) %>%
  summarise(n = n(), bp = sum(end - start), .groups = "keep") %>%
  group_by(scaff, seqOnt) %>%
  arrange(desc(bp)) %>%
  head(50) 
```

### Make a list of families with at least # members

Five thousand seems like a reasonable cutoff? The bigFams list can be used for plots of both H1 and H2, maybe?

```{r listOfFamiliesWith5kMembers}
cutoff = 5000

H1bigFams <- H1TEs %>%
  group_by(family) %>%
  summarize(n = n()) %>%
  arrange( desc(n) ) %>%
  filter(n > cutoff) %>%
  pull(family) #25k
  #head(100) #132 864 - 2 658
H2bigFams <- H2TEs %>%
  group_by(family) %>%
  summarize(n = n()) %>%
  arrange( desc(n) ) %>% 
  filter(n > cutoff) %>%
  pull(family) #30k 
  #head(100) #38 975 - 2 424 

bigFamsAll <- append(H1bigFams, H2bigFams)
bigFamsSorted <- sort(bigFamsAll)

bigFams <- unique(bigFamsSorted)

rm(cutoff, H1bigFams, H2bigFams, bigFamsAll, bigFamsSorted)
```


### Combine family data for H1 and H2

Because H1 and H2 are only really differentiated on the sex chromosomes (we're not confident that the autosomes really belong to one haplotype or the other), I'm going to keep only the TEs that appear on the sex chromosomes.

```{r combineFamilyData}
H1fams <- H1TEs %>%
  filter(scaff == "X") %>%
  group_by(seqOnt, family) %>%
  summarise(n = n(), bp = sum(end - start), .groups = "keep") %>%
  group_by(seqOnt)

H2fams <- H2TEs %>%
  filter(scaff == "Y1" | scaff == "Y2") %>%
  group_by(seqOnt, family) %>%
  summarise(n = n(), bp = sum(end - start), .groups = "keep") %>%
  group_by(seqOnt) 

# Divide the # of bp occupied by the total # of bp available on the chromosome
Xsize = scaffLens$H1Len[4]
Ysize = scaffLens$H2Len[4] + scaffLens$H2Len[5]



# Keeping bpH1 and bpH2 wide
famTally <- full_join(H1fams, H2fams, by = join_by(family, seqOnt),
                      suffix = c("H1", "H2")) %>%
  replace_na(list(nH1 = 0, bpH1 = 0, nH2 = 0, bpH2 = 0)) %>% #if missing from a haplotype, the # present is 0
  mutate(bp = bpH1 + bpH2, n = nH1 + nH2,
         fracH1 = bpH1 / Xsize, fracH2 = bpH2 / Ysize) %>%
  #select(!c(bpH1, bpH2)) %>% #15 193
  filter(n > 1) #3 738





# Transform the data so haplotype is a factor
famFacTally <- full_join(H1fams, H2fams, by = join_by(family, seqOnt),
                      suffix = c("H1", "H2")) %>%
  replace_na(list(nH1 = 0, bpH1 = 0, nH2 = 0, bpH2 = 0)) %>% #if missing from a haplotype, the # present is 0
  mutate(count = nH1 + nH2, 
         total = bpH1 + bpH2,
         fracH1 = bpH1 / Xsize, fracH2 = bpH2 / Ysize) %>%
  select(!c(nH1, nH2, bpH1, bpH2)) %>%
  filter(count > 1) %>%
  # pivot_longer(c(bpH1, bpH2, nH1, nH2), 
  #              names_to = c(".value", "haplotype"), 
  #              names_pattern = "(bp|n)(H.)") %>%
  pivot_longer(c(fracH1, fracH2), 
               names_to = "haplotype", values_to = "fraction")


rm(H1fams, H2fams, Xsize, Ysize)
```


## Windowing

```{r cleanH1, message=FALSE}
H1EDTA <- H1EDTAraw %>%
  filter(scaff %in% H1scaffs) %>%
  filterTEs %>%
  window %>%
  summarizeTEs %>%
  renameTEs

H1Intact <- H1IntactRaw %>%
  filter(scaff %in% H1scaffs) %>%
  filterTEs %>%
  window %>%
  summarizeTEs %>%
  renameTEs

H1Genes <- H1GenesRaw %>%
  filter(scaff %in% H1scaffs) %>%
  filterGenes %>%
  window %>%
  summarizeGenes

H1Lifted <- H1LiftedRaw %>%
  filter(scaff %in% H1scaffs) %>%
  filter(seqOnt == "gene") %>%
  filterGenes %>%
  window %>%
  summarizeGenes

# H1CDS <- H1MakerRaw %>%
#   filterMAKER %>%
#   group_by(scaff) %>%
#   window %>%
#   summarizeGenes
```

```{r cleanH2, message=FALSE}
H2EDTA <- H2EDTAraw %>%
  filter(scaff %in% H2scaffs) %>%
  filterTEs %>%
  window %>%
  summarizeTEs %>%
  renameTEs

H2Intact <- H2IntactRaw %>%
  filter(scaff %in% H2scaffs) %>%
  filterTEs %>%
  window %>%
  summarizeTEs %>%
  renameTEs

H2Genes <- H2GenesRaw %>%
  filter(scaff %in% H2scaffs) %>%
  filterGenes %>%
  window %>%
  summarizeGenes

H2Lifted <- H2LiftedRaw %>%
  filter(scaff %in% H2scaffs) %>%
  filter(seqOnt == "gene") %>%
  filterGenes %>%
  window %>%
  summarizeGenes

# H2CDS <- H2MakerRaw %>%
#   filterMAKER %>%
#   group_by(scaff) %>%
#   window %>%
#   summarizeGenes
```

```{r deleteRawFiles}
#rm(NCfemEDTAraw, NCfemIntactRaw, NCfemGenesRaw)
rm(H1EDTAraw, H1IntactRaw, H1GenesRaw, H1LiftedRaw)
rm(H2EDTAraw, H2IntactRaw, H2GenesRaw, H2LiftedRaw)
```

## Divide up sex chromosomes

### TEs

Split the EDTA annotations for X and Y into PAR1/2, old, and new based on window number (approximate position). Join the H1 and H2 sets of TE bp per window into *allSummedTEs*.

```{r TEsInSexRegions}
XsummedTEs <- H1EDTA %>%
  filter(scaff == "X") %>%
  splitX %>%
  group_by(seqOnt, region) %>%
  summarize(summedbp = sum(bp), .groups = "keep") %>%
  mutate(X = case_when(
    region == "PAR1"   ~ summedbp/71000000,
    region == "Old"    ~ summedbp/(261000000-71000000),
    region == "New"    ~ summedbp/(367000000-261000000),
    region == "PAR2"   ~ summedbp/(483000000-367000000)
  ))

Y1summedTEs <- H2EDTA %>%
  filter(scaff == "Y1") %>%
  splitY1 %>%
  group_by(seqOnt, region) %>%
  summarize(summedbp = sum(bp), .groups = "keep") %>%
  mutate(Y1 = case_when(
    region == "New"   ~ summedbp/64000000,
    region == "Old"    ~ summedbp/(272000000-64000000),
    region == "PAR1"    ~ summedbp/(343000000-272000000)
  ))

Y2summedTEs <- H2EDTA %>%
  filter(scaff == "Y2") %>%
  splitY2 %>%
  group_by(seqOnt, region) %>%
  summarize(summedbp = sum(bp), .groups = "keep") %>%
  mutate(Y2 = case_when(
    region == "PAR2"   ~ summedbp/121000000,
    region == "New"    ~ summedbp/(159000000-121000000),
    region == "Old"    ~ summedbp/(348000000-159000000)
  ))

allSummedTEs <- full_join(XsummedTEs, Y1summedTEs, 
                          by = join_by(seqOnt, region)) %>%
  full_join(Y2summedTEs, by = join_by(seqOnt, region)) %>%
  pivot_longer(cols = c(X, Y1, Y2), 
               names_to = "Chromosome", values_to = "bp")

# Clean up intermediate files
rm(XsummedTEs, Y1summedTEs, Y2summedTEs)
```

### Genes

```{r genesInSexRegions}
XsummedGenes <- H1Genes %>%
  filter(scaff == "X") %>%
  splitX %>%
  group_by(region) %>%
  summarize(summedbp = sum(bp), .groups = "keep") %>%
  mutate(X = case_when(
    region == "PAR1"   ~ summedbp/71000000,
    region == "Old"    ~ summedbp/(261000000-71000000),
    region == "New"    ~ summedbp/(367000000-261000000),
    region == "PAR2"   ~ summedbp/(483000000-367000000)
  )) %>%
  select(region, X)

Y1summedGenes <- H2Genes %>%
  filter(scaff == "Y1") %>%
  splitY1 %>%
  group_by(region) %>%
  summarize(summedbp = sum(bp), .groups = "keep") %>%
  mutate(Y1 = case_when(
    region == "New"   ~ summedbp/64000000,
    region == "Old"    ~ summedbp/(272000000-64000000),
    region == "PAR1"    ~ summedbp/(343000000-272000000)
  )) %>%
  select(region, Y1)

Y2summedGenes <- H2Genes %>%
  filter(scaff == "Y2") %>%
  splitY2 %>%
  group_by(region) %>%
  summarize(summedbp = sum(bp), .groups = "keep") %>%
  mutate(Y2 = case_when(
    region == "PAR2"   ~ summedbp/121000000,
    region == "New"    ~ summedbp/(159000000-121000000),
    region == "Old"    ~ summedbp/(348000000-159000000)
  )) %>%
  select(region, Y2)

allSummedGenes <- full_join(XsummedGenes, Y1summedGenes, 
                            by = join_by(region)) %>%
  full_join(Y2summedGenes, by = join_by(region)) %>%
  pivot_longer(cols = c(X, Y1, Y2), 
               names_to = "Chromosome", values_to = "bp")

# Clean up intermediate files
rm(XsummedGenes, Y1summedGenes, Y2summedGenes)
```

# Visualise

## Functions

```{r functionsForWindowedPlotting}
plotWindowedTEs <- function(df, name) {
  ggplot(df, aes(winnum, n, colour = seqOnt), size = 0.3) +
    geom_line() +
    pubTheme +
    labs(x = paste(name, "position (1Mb windows)"),
         y = "Number of TEs", colour = "TE Type") +
    facet_grid(vars(scaff))
}

plotElements <- function(TEs, genes, name) {
  ggplot() +
  geom_line(data = TEs, aes(winnum, amt, colour=seqOnt), linewidth=0.3) +
  geom_line(data = genes, aes(winnum,amt)) +
  pubTheme +
  labs(x = paste(name, "position (1Mb windows)"), 
       y = paste(name, "Fraction of window"), colour="TE Type") +
  facet_grid(vars(scaff), scale="free_x", space="free")
}

# plotFacetedTEs <- function(df, name) {
#   ggplot(df, aes(winnum, n), size = 0.3) +
#   geom_line() +
#   pubTheme +
#   labs(x = paste(name, "position (1Mb windows)"),
#        y = "Number of TEs", colour = "TE Type") +
#   facet_grid(vars(seqOnt), vars(scaff), space = "free", scale = "free")
# }
```

```{r functionPlotFamilies}
plotFamHisto <- function(df, name) {
  ggplot(df, aes(start, fill = seqOnt)) +
    geom_histogram(binwidth = 1000000) +
    pubTheme +
    labs(x = paste(name,"start position (1Mb bins)"), 
         y = "Number of TEs in this bin that belong to a large family",
         fill = "TE Type") +
    facet_grid(vars(scaff))
}

plotFamBoxes <- function(df, name) {
  ggplot(df, aes(scaff, logN, colour = seqOnt)) +
  geom_boxplot() +
  pubTheme + 
  labs(x = paste(name, "Scaffold"), 
       y = "Log Number of family members", colour = "TE Type")
}

```

These bar plot functions are a bit trickier, just because there's stuff to control for chromosome length. The variables for each chromosome's length are set in the Import section.

```{r functionsForBarPlots}
# Control number of elements for Scaffold Length
controlH1Counts <- function(df) {
  df %>%
    group_by(seqOnt, scaff) %>%
    tally() %>%
    mutate( num = case_when(
      scaff == "A1"     ~ n/scaffLens$H1Len[1],
      scaff == "A2"     ~ n/scaffLens$H1Len[2],
      scaff == "A4"     ~ n/scaffLens$H1Len[3],
      scaff == "X"      ~ n/scaffLens$H1Len[4]
    )) 
}
controlH2Counts <- function(df) {
  df %>%
    group_by(seqOnt, scaff) %>%
    tally() %>%
    mutate( num = case_when(
      scaff == "A1"   ~ n/scaffLens$H2Len[1],
      scaff == "A2"   ~ n/scaffLens$H2Len[2],
      scaff == "A4"   ~ n/scaffLens$H2Len[3],
      scaff == "Y1" ~ n/scaffLens$H2Len[4],
      scaff == "Y2" ~ n/scaffLens$H2Len[5]
    ))
}

# Divide length of elements by Scaffold Length
controlH1Len <- function(df) {
  df %>%
    group_by(seqOnt, scaff) %>%
    summarise(n = sum(bp)) %>%
    mutate( num = case_when(
      scaff == "A1"     ~ n/scaffLens$H1Len[1],
      scaff == "A2"     ~ n/scaffLens$H1Len[2],
      scaff == "A4"     ~ n/scaffLens$H1Len[3],
      scaff == "X"      ~ n/scaffLens$H1Len[4]
    )) 
}
controlH2Len <- function(df) {
  df %>%
    group_by(seqOnt, scaff) %>%
    summarise(n = sum(bp)) %>%
    mutate( num = case_when(
      scaff == "A1"   ~ n/scaffLens$H2Len[1],
      scaff == "A2"   ~ n/scaffLens$H2Len[2],
      scaff == "A4"   ~ n/scaffLens$H2Len[3],
      scaff == "Y1" ~ n/scaffLens$H2Len[4],
      scaff == "Y2" ~ n/scaffLens$H2Len[5]
    ))
}

# Make bar plot 
plotCols <- function(df, name, legPos) {
  ggplot(df, aes(x = seqOnt, y = num, fill = scaff)) +
  geom_col(position = position_dodge()) +
  pubTheme + theme(legend.position = legPos) +
  labs(x = "TE Type", y = "TEs divided by scaffold length", 
       fill = paste(name, "scaffold"))
}

# Full chromosomal proportions
plotProportions <- function(TEs, Genes, name, legPos) {
  TEsums <- TEs %>%
    group_by(scaff) %>%
    summarize(TEs = sum(bp))
  geneSums <- Genes %>%
    group_by(scaff) %>%
    summarize(Genes = sum(bp))
  df <- left_join(TEsums, geneSums) %>%
    pivot_longer(cols = c(TEs, Genes), 
                 names_to = "element", values_to = "bp")
  
  ggplot(df, aes(x = scaff, y = bp, fill = element)) +
    geom_col() +
    pubTheme + theme(legend.position = legPos) +
    labs(x = paste(name, "chromosome"), y = "Base pairs", 
         fill = "Element type")
}
```


## TE barplots {.tabset}

### Frequency and Count per scaffold

Plot the sum of TE bp divided by the length of the scaffold they're found on.

```{r fractionCols, message=FALSE}
H1EDTA %>%
  controlH1Len() %>%
  plotCols("H1", c(0.25, 0.7))

H2EDTA %>%
  controlH2Len() %>%
  plotCols("H2", c(0.25, 0.7))

# H1Intact %>%
#   controlH1Len() %>%
#   plotCols("H1", c(0.25, 0.7))
# 
# H2Intact %>%
#   controlH2Len() %>%
#   plotCols("H2", c(0.25, 0.7))
```

The panEDTA data has less of a strong pattern than the original annotation. 

```{r countCols, message=FALSE}
H1EDTA %>%
  controlH1Counts() %>%
  plotCols("H1", c(0.25, 0.7))

H2EDTA %>%
  controlH2Counts() %>%
  plotCols("H2", c(0.25, 0.7))

# H1Intact %>%
#   controlH1Counts() %>%
#   plotCols("H1", c(0.25, 0.7))
# 
# H2Intact %>%
#   controlH2Counts() %>%
#   plotCols("H2", c(0.25, 0.7))
```

The untransformed count data is hard to parse (and generally just doesn't make sense).

### Sex chromosome proportions

The first plot (with region on the x-axis) feels like the best way to visualise the comparisons we're interested in. 

```{r plotSummedTEs, warning=FALSE}
ggplot(allSummedTEs, aes(x = region, y = bp, fill = Chromosome)) +
  geom_col(position = position_dodge()) +
  pubTheme + scale_fill_manual(values = pubColours) +
  labs(x = "Region", y = "Fraction of region occupied by TEs", 
       fill = "Chromosome") #+
  #facet_grid(vars(seqOnt))

ggplot(allSummedTEs, aes(x = Chromosome, y = bp, fill = region)) +
  geom_col(position = position_dodge()) +
  pubTheme + scale_fill_manual(values = pubColours) +
  labs(x = "Chromosome", y = "Fraction of region occupied by TEs", 
       fill = "Region")

allSummedTEs %>%
  #filter(seqOnt == "DTM") %>%
  ggplot(aes(x = seqOnt, y = bp, fill = Chromosome)) +
  geom_col(position = position_dodge()) +
  pubTheme + scale_fill_manual(values = pubColours) +
  theme(legend.position = c(0.2,0.8)) +
  labs(x = "Region", y = "Fraction of region occupied by TEs", 
       fill = "Chromosome") +
  facet_grid(vars(region))
```

```{r plotSummedGenes}
ggplot(allSummedGenes, aes(x = region, y = bp, fill = Chromosome)) +
  geom_col(position = position_dodge()) + 
  theme(legend.position = c(0.13, 0.78)) +
  pubTheme + scale_fill_manual(values = pubColours) +
  labs(x = "Region", y = "Fraction of region occupied by Genes", 
       fill = "Chromosome")
```


## Chromosomal landscape {.tabset}

Fraction of window annotated with Genes, TEs

### TE Frequencies

```{r plotElementFrequencies}
plotElements(H1EDTA, H1Genes, "H1")
plotElements(H2EDTA, H2Genes, "H2")


H2EDTA %>%
  filter(scaff == "Y1" | scaff == "Y2") %>%
  ggplot() +
  geom_line(data = H1EDTA, aes(winnum, amt, colour=seqOnt), 
            linewidth = 0.3) +
  geom_line(aes(winnum, amt, colour=seqOnt), linewidth = 0.3) +
  geom_line(data = H1Genes, aes(winnum,amt)) +
  geom_line(data = filter(H2Genes, scaff == "Y1" | scaff == "Y2"),
            aes(winnum,amt)) +
  pubTheme +
  labs(x = "Position (1Mb windows)", y = "Fraction of window", 
       colour="TE Type") +
  facet_grid(vars(scaff), scale="free_x", space="free")
```

### TE counts

```{r plotTEcounts}
plotWindowedTEs(H1EDTA, "H1")
plotWindowedTEs(H2EDTA, "H2")
```


## TE families {.tabset}

### Visualising large families

```{r bigFamHistogram, exec=FALSE}
# Count
H1TEs %>%
  group_by(family) %>%
  summarise(n = n()) %>%
  ggplot() +
  geom_histogram(aes(n), bins=100) +
  ylim(0,100)

# Size
H1TEs %>%
  group_by(family) %>%
  summarise(size = sum(end - start)) %>%
  ggplot() +
  geom_histogram(aes(size), bins=100) +
  ylim(0,100)
```

Plotting the 50 biggest families, to see what type they are. 

```{r barplot50biggestFamSizes}
H1TEs %>%
  group_by(family, seqOnt) %>%
  summarise(size = sum(end - start) / 1000000, 
            n = n(), .groups = "keep") %>%
  filter(seqOnt != "low_complexity") %>%
  arrange(desc(size)) %>%
  head(50) %>%
  ggplot(aes(x = reorder(family, -size), y = size)) +
  geom_col(aes(fill = seqOnt)) + 
  geom_text(aes(label = n), size = 3,  angle = 85) +
  pubTheme + theme(legend.position = c(0.8, 0.6)) +
  labs(x = "50 largest TE families in H1", 
       y = "Size (1Mb)", fill = "TE Type")

H2TEs %>%
  group_by(family, seqOnt) %>%
  summarise(size = sum(end - start) / 1000000, 
            n = n(), .groups = "keep") %>%
  arrange(desc(size)) %>%
  head(50) %>%
  ggplot(aes(x = reorder(family, -size), y = size)) +
  geom_col(aes(fill = seqOnt)) + 
  geom_text(aes(label = n, y = size + 4), size = 3, angle = 85) +
  pubTheme + theme(legend.position = c(0.8, 0.6)) +
  labs(x = "50 largest TE families in H2", 
       y = "Size (1Mb)", fill = "TE Type")




# Count
H1TEs %>%
  group_by(family, seqOnt) %>%
  summarise(size = sum(end - start) / 1000000, 
            n = n(), .groups = "keep") %>%
  arrange(desc(n)) %>%
  head(50) %>%
  ggplot(aes(x = reorder(family, -n), y = n)) +
  geom_col(aes(fill = seqOnt)) + 
  geom_text(aes(label = size, y = n + 4), size = 3, angle = 85) +
  pubTheme + theme(legend.position = c(0.8, 0.6)) +
  labs(x = "50 largest TE families in H1", 
       y = "Copy number", fill = "TE Type")

H2TEs %>%
  group_by(family, seqOnt) %>%
  summarise(size = sum(end - start) / 1000000, 
            n = n(), .groups = "keep") %>%
  arrange(desc(n)) %>%
  head(50) %>%
  ggplot(aes(x = reorder(family, -n), y = n)) +
  geom_col(aes(fill = seqOnt)) + 
  geom_text(aes(label = size, y = n + 4), size = 3, angle = 85) +
  pubTheme + theme(legend.position = c(0.8, 0.6)) +
  labs(x = "50 largest TE families in H2", 
       y = "Copy number", fill = "TE Type")
```

Some big families!! Not drastically larger than with the old data set, but a bit!

### Family size differences across sex chromosomes

Bar plots of the largest families

```{r barplotBiggestFams}
# # Using haplotype as a factor (which is probably cleaner?)
famFacTally %>%
  ggplot(aes(x = reorder(family, -total), y = total / 1000000)) +
  geom_col(aes(fill = haplotype, colour = seqOnt)) +
  scale_fill_manual(values = c("black", "darkgrey")) +
  #geom_text(aes(y = bp / 1000000 + 8, label = n), size = 3,  angle = 85) +
  pubTheme + theme(legend.position = c(0.88, 0.61)) +
  labs(x = "50 largest TE families across haplotypes",
       y = "Size (Mb occupied)", colour = "TE Type", fill = "Haplotype")

# Using points to differentiate haplotypes
famTally %>%
  # Keep only the largest families
  arrange(desc(bp)) %>%
  head(50) %>%
  ggplot(aes(x = reorder(family, -bp), y = bp / 1000000)) +
  # The total across both haplotypes
  geom_col(aes(fill = seqOnt)) +
  # Label columns with copy number across haplotypes
  geom_text(aes(y = bp / 1000000 + 5, label = n), size = 3,  angle = 85) +
  # Use points to differentiate haplotypes
  geom_point(aes(y = bpH1 / 1000000), shape = 1, size = 2, stroke = 1.2) +
  geom_point(aes(y = bpH2 / 1000000), shape = 2, size = 2, stroke = 1.2) +
  pubTheme + theme(legend.position = c(0.88, 0.67)) +
  labs(x = "50 largest TE families across haplotypes",
       y = "Size (Mb occupied)", fill = "TE Type")
```

### Scatterplot of family size on Haplotype 1 and Haplotype 2

```{r scatterplotFamAmounts}
# Use a slope that represents the size difference between the chromosomes
slopey = (scaffLens$H2Len[4] + scaffLens$H2Len[5]) / scaffLens$H1Len[4]

famTally %>%
  ggplot(aes(x = fracH1, y = fracH2, colour = seqOnt)) +
  geom_abline(slope = 1) +
  # The total across both haplotypes
  geom_point(aes(size = n)) +
  #geom_text(aes(y = bp / 1000000 + 8, label = n), size = 3,  angle = 85) +
  pubTheme + #xlim(0, 50) + ylim(0, 63) + #theme(legend.position = c(0.88, 0.67)) +
  labs(x = "Fraction of the X occupied", size = "Number",
       y = "Fraction of Y1 and Y2 occupied", colour = "TE Type")

famTally %>%
  ggplot(aes(x = nH1, y = nH2, colour = seqOnt)) +
  geom_abline(slope = slopey) +
  # The total across both haplotypes
  geom_point(aes(size = bp / winsize)) +
  #geom_text(aes(y = bp / 1000000 + 8, label = n), size = 3,  angle = 85) +
  pubTheme + #theme(legend.position = "none") +
  labs(x = "Copy number on the X", size = "Size (Mb)",
       y = "Copy number on Y1 and Y2", colour = "TE Type")

rm(slopey)
```

Accounting for total size 

```{r famsFacetedByTEtype}
# Use a slope that represents the size difference between the chromosomes
slopey = (scaffLens$H2Len[4] + scaffLens$H2Len[5]) / scaffLens$H1Len[4]

famTally %>%
  ggplot(aes(x = fracH1, y = fracH2, colour = seqOnt)) +
  geom_abline(slope = 1) +
  # The total across both haplotypes
  geom_point(aes(size = n)) +
  #geom_text(aes(y = bp / 1000000 + 8, label = n), size = 3,  angle = 85) +
  pubTheme + theme(legend.position = "none") +
  labs(x = "Fraction of the X occupied", size = "Number",
       y = "Size on Y1 and Y2 (Mb occupied)", colour = "TE Type") +
  facet_wrap(vars(seqOnt), scales = "free")

famTally %>%
  ggplot(aes(x = nH1, y = nH2, colour = seqOnt)) +
  geom_abline(slope = slopey) +
  # The total across both haplotypes
  geom_point(aes(size = bp / winsize)) +
  #geom_text(aes(y = bp / 1000000 + 8, label = n), size = 3,  angle = 85) +
  pubTheme + theme(legend.position = "none") +
  labs(x = "Copy number on the X", size = "Size (Mb)",
       y = "Copy number on Y1 and Y2", colour = "TE Type") +
  facet_wrap(vars(seqOnt), scales = "free")

rm(slopey)
```


### Boxplots

Plotting the log copy number of families, grouped by chromosome and TE type. Because of the extremely large values (and discrepancy between the. majority-low values and extreme outliers on the high end), the log transform just helps to visualise things.

```{r boxplotsOfFamilies}
H1TEs %>%
  #filter(scaff == "X" | scaff == "A1") %>%
  group_by(scaff, seqOnt, family) %>%
  summarise(n = n(), bp = sum(end - start), .groups = "keep") %>%
  group_by(scaff, seqOnt) %>%
  filter(n != 1) %>% #no singletons considered for "families"
  mutate(logN = log(n)) %>%
  plotFamBoxes("H1")

H2TEs %>%
  #filter(scaff == "Y1" | scaff == "Y2") %>%
  group_by(scaff, seqOnt, family) %>%
  summarise(n = n(), bp = sum(end - start), .groups = "keep") %>%
  group_by(scaff, seqOnt) %>%
  filter(n != 1) %>% #no singletons considered for "families"
  mutate(logN = log(n)) %>%
  plotFamBoxes("H2")
```

### Chromosomal Landscape

Make a histogram of TEs belonging to families with at least 5k members.

```{r visualLargeFamilies}
h1biggies <- H1TEs %>%
  filter(scaff == "X" | scaff == "A1") %>%
  filter(family %in% bigFams) #%>%

h2biggies <- H2TEs %>%
  filter(scaff == "Y1" | scaff == "Y2") %>%
  filter(family %in% bigFams) #%>%
  # plotFamHisto("H2") +
  # theme(legend.position = c(0.87,0.45))

h1biggies %>%
  bind_rows(h2biggies) %>%
  ggplot(aes(start, fill = seqOnt)) +
  geom_histogram(binwidth = 1000000) +
  pubTheme + theme(legend.position = c(0.9, 0.35)) +
  labs(x = "Start position (1Mb bins)", 
       y = "Number of TEs in this bin that belong to a large family",
       fill = "TE Type") +
  facet_grid(vars(scaff), scales = "free", space = "free")

rm(h1biggies, h2biggies)
```

The plot shows there has been extreme proliferation of TEs on Y1 and fairly significant accumulation on Y2 as well. 

### Biggest by superfamily

Identify the five biggest families (by copy number) in the most abundant superfamilies: RLC, DTM, RLX, DTH, and RLG

```{r superFamBiggies}
bigSuperFams <- famTally %>%
  filter(seqOnt == "RLC" | seqOnt == "DTM" | seqOnt == "RLX" |
         seqOnt == "DTH" | seqOnt == "RLG") %>%
  group_by(seqOnt) %>%
  slice_max(order_by = n, n = 5) %>%
  pull(family)

h1biggies <- H1TEs %>%
  filter(scaff == "X" | scaff == "A1") %>%
  filter(family %in% bigSuperFams) #%>%

h2biggies <- H2TEs %>%
  filter(scaff == "Y1" | scaff == "Y2") %>%
  filter(family %in% bigSuperFams)



# Starting to plot the families

h1biggies %>%
  bind_rows(h2biggies) %>%
  filter(seqOnt == "RLC") %>%
  ggplot(aes(start, fill = seqOnt)) +
  geom_histogram(binwidth = 1000000) +
  pubTheme + theme(legend.position = c(0.87,0.3)) +
  scale_fill_manual(values = pubColours[1]) +
  labs(x = "Start position (1Mb bins)", 
       y = "Number of TEs that belong to the 5 biggest families",
       fill = "TE Type") +
  facet_grid(vars(scaff))#, scales = "free")#, strip.position = "bottom")

h1biggies %>%
  bind_rows(h2biggies) %>%
  filter(seqOnt == "RLG") %>%
  ggplot(aes(start, fill = seqOnt)) +
  geom_histogram(binwidth = 1000000) +
  pubTheme + theme(legend.position = c(0.87,0.3)) +
  scale_fill_manual(values = pubColours[2]) +
  labs(x = "Start position (1Mb bins)", 
       y = "Number of TEs that belong to the 5 biggest families",
       fill = "TE Type") +
  facet_grid(vars(scaff))

h1biggies %>%
  bind_rows(h2biggies) %>%
  filter(seqOnt == "RLX") %>%
  ggplot(aes(start, fill = seqOnt)) +
  geom_histogram(binwidth = 1000000) +
  pubTheme + theme(legend.position = c(0.87,0.3)) +
  scale_fill_manual(values = pubColours[3]) +
  labs(x = "Start position (1Mb bins)", 
       y = "Number of TEs that belong to the 5 biggest families",
       fill = "TE Type") +
  facet_grid(vars(scaff))

h1biggies %>%
  bind_rows(h2biggies) %>%
  filter(seqOnt == "DTM") %>%
  ggplot(aes(start, fill = seqOnt)) +
  geom_histogram(binwidth = 1000000) +
  pubTheme + theme(legend.position = c(0.87,0.3)) +
  scale_fill_manual(values = pubColours[4]) +
  labs(x = "Start position (1Mb bins)", 
       y = "Number of TEs that belong to the 5 biggest families",
       fill = "TE Type") +
  facet_grid(vars(scaff))

h1biggies %>%
  bind_rows(h2biggies) %>%
  filter(seqOnt == "DTH") %>%
  ggplot(aes(start, fill = seqOnt)) +
  geom_histogram(binwidth = 1000000) +
  pubTheme + theme(legend.position = c(0.87,0.3)) +
  scale_fill_manual(values = pubColours[5]) +
  labs(x = "Start position (1Mb bins)", 
       y = "Number of TEs that belong to the 5 biggest families",
       fill = "TE Type") +
  facet_grid(vars(scaff))

rm(h1biggies, h2biggies, bigSuperFams)
```


# Statistics

## Are TE families larger on the Ys than on the X?

I do need to consider the fact that the Ys are larger than the X.

```{r statsForFamDiffs}
head(famFacTally, 10)
# X length: 483481970, Y: 691272551
#scaffLens$H1Len[4] / (scaffLens$H2Len[4] + scaffLens$H2Len[5])

t.test(fraction ~ haplotype, data = famFacTally, paired = TRUE)
```


# Testing things

## Gene sets

Green is the lifted over data set (from H2 genes) and blue is lifted over from H1

```{r geneSets}
ggplot(H1Genes, aes(winnum, n)) +
  geom_line() +
  geom_line(data = H1Lifted, aes(winnum, n), colour = pubColours[2]) +
  pubTheme +
  labs(x = "H1 Position (1Mb windows)", y = "Number of Genes") +
  facet_grid(vars(scaff))

ggplot(H2Genes, aes(winnum, n)) +
  geom_line() +
  geom_line(data = H2Lifted, aes(winnum, n), colour = pubColours[1]) +
  pubTheme +
  labs(x = "H2 Position (1Mb windows)", y = "Number of Genes") +
  facet_grid(vars(scaff))



ggplot(H1Genes, aes(winnum, bp)) +
  geom_line() +
  geom_line(data = H1Lifted, aes(winnum, bp), colour = pubColours[2]) +
  pubTheme +
  labs(x = "H1 Position (1Mb windows)", y = "Number of bp") +
  facet_grid(vars(scaff))

ggplot(H2Genes, aes(winnum, bp)) +
  geom_line() +
  geom_line(data = H2Lifted, aes(winnum, bp), colour = pubColours[1]) +
  pubTheme +
  labs(x = "H2 Position (1Mb windows)", y = "Number of bp") +
  facet_grid(vars(scaff))



ggplot(H1Genes, aes(winnum, amt)) +
  geom_line() +
  geom_line(data = H1Lifted, aes(winnum, amt), colour = pubColours[2]) +
  pubTheme +
  labs(x = "H1 Position (1Mb windows)", y = "Fraction of window") +
  facet_grid(vars(scaff))

ggplot(H2Genes, aes(winnum, amt)) +
  geom_line() +
  geom_line(data = H2Lifted, aes(winnum, amt), colour = pubColours[1]) +
  pubTheme +
  labs(x = "H2 Position (1Mb windows)", y = "Fraction of window") +
  facet_grid(vars(scaff))





ggplot(H1Genes, aes(winnum, amt)) +
  geom_line(data = H1Genes, colour = "forestgreen") +
  geom_line(data = H1Lifted, colour = pubColours[2]) +
  geom_line(data = H2Genes, colour = "navy") +
  geom_line(data = H2Lifted, colour = pubColours[1]) +
  pubTheme +
  labs(x = "H1 Position (1Mb windows)", y = "Fraction of window") +
  facet_grid(vars(scaff))

ggplot(H1Genes, aes(winnum, n)) +
  geom_line(data = H1Genes, colour = "forestgreen") +
  geom_line(data = H1Lifted, colour = pubColours[2]) +
  geom_line(data = H2Genes, colour = "navy") +
  geom_line(data = H2Lifted, colour = pubColours[1]) +
  pubTheme +
  labs(x = "H1 Position (1Mb windows)", y = "Number of genes") +
  facet_grid(vars(scaff))
```


## CDS instead of gene?

```{r testingGeneFiltering, eval=FALSE}
H1MakerRaw %>%
  filter(type == "CDS") %>%
  group_by(scaff) %>%
  window %>%
  summarizeGenes %>%
  ggplot() +
  geom_line(data = H1Genes, aes(winnum, amt), colour = pubColours[1], size = 1.1) +
  geom_line(aes(winnum, amt), colour = pubColours[3], size = 1.2) +
  geom_line(data = H1CDS, aes(winnum, amt), colour = pubColours[2], size = 1.1) +
  pubTheme +
  labs(x = "H1 position (1Mb windows)",
       y = "CDS frequency") +
  facet_grid(vars(scaff), scale="free_x", space="free")
```

```{r CDSvsGene, eval=FALSE}
# Plot counts
H1CDS %>%
  ggplot() +
  geom_line(data = H1Genes, aes(winnum,  n), colour = pubColours[1], size = 1.1) +
  geom_line(aes(winnum, n), colour = pubColours[2], size = 1.1) +
  pubTheme +
  labs(x = "H1 position (1Mb windows)", 
       y = "Element count") +
  facet_grid(vars(scaff), scale="free_x", space="free")

H2CDS %>%
  ggplot() +
  geom_line(data = H2Genes, aes(winnum, n), colour = pubColours[1], size = 1.1) +
  geom_line(aes(winnum, n), colour = pubColours[2], size = 1.1) +
  pubTheme +
  labs(x = "H2 position (1Mb windows)", 
       y = "Element count") +
  facet_grid(vars(scaff), scale="free_x", space="free")




# Plot densities (equivalent to bp)
H1CDS %>%
  ggplot() +
  geom_line(data = H1Genes, aes(winnum, amt), colour = pubColours[1]) +
  geom_line(aes(winnum, amt), colour = pubColours[2]) +
  pubTheme +
  labs(x = "H1 position (1Mb windows)", 
       y = "Fraction of window") +
  facet_grid(vars(scaff), scale="free_x", space="free")

H2CDS %>%
  ggplot() +
  geom_line(data = H2Genes, aes(winnum, amt), colour = pubColours[1]) +
  geom_line(aes(winnum, amt), colour = pubColours[2]) +
  pubTheme +
  labs(x = "H2 position (1Mb windows)", 
       y = "Fraction of window") +
  facet_grid(vars(scaff), scale="free_x", space="free")
```

