---
title: "Relaxed Selection against TE Insertions"
author: "Zoë Humphries"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    code_folding: hide
    default_style: dark
    toc_depth: 6
    fig_caption: yes
---

```{r setup, include=FALSE}
#echo=FALSE, results=FALSE, message=FALSE, warning=FALSE
# Set the root directory
#knitr::opts_knit$set(root.dir = "~/OneDrive/GradSchool/3_Analysis/2022NewRefs")
knitr::opts_knit$set(root.dir = "~/2022NewRefs")

# Import libraries
library("tidyverse") #so much love
library("patchwork") #for arranging plots into figures

# Colouring & themes
pubTheme <-
  theme(title = element_text(size=10), #usually 14
        text = element_text(size=8),
        plot.background = element_rect(fill="lightgrey"),
        #plot.title = element_text(hjust = 0.5), #not sure yet if I want this
        strip.background = element_rect(linetype=0,linewidth=8,
                                        fill="grey"), #facet boxes
        strip.text = element_text(size=10), #usually 11
        legend.background = element_rect(fill="grey"),
        legend.text = element_text(size=8),
        legend.title = element_text(size=10),
        panel.background = element_rect(fill="#FFFFFF"), #plot body
        panel.grid = element_blank() #plot body pt2
  )
#thesisColours <- c("#d84638","#65bc5c","#01a5bc","#516fc8","#373f51") #dark
pubColours <- c("#2d9da6","#99b700","#8a53b6","#bf4824","#152944") #for white paper
```

# Data Overview {.tabset}

## Input 

The raw MAKER data set was filtered for one-to-one syntenic orthlogs present in haplotype 1 and haplotype 2 (data generated by Bianca's GENESPACE run). This reduced the set of genes from ~27k to ~16k (as was output from GENESPACE).

Output from bedtools coverage:
- chromosome
- region (upstream/downstream of gene, intron, or exon) start
- region end
- number of TEs that overlapped a region by at least one bp
- number of bases in the region with coverage from a TE
- region length
- fraction of bases in the region with coverage from a TE

## Created

windowAndMerge output variables:
- n :: the number of elements in the window (upNum=downNum unless at the beginning/end of the chromosome)
- sum :: the number of TEs in elements in the window
- len :: the total number of TE bp in regions within the window
- avg :: the mean frequency of TEs in elements in the window

# Import & Analysis {.tabset}

## Functions and variables

Define window size (1Mb probably), column names, and chromosome names.

```{r variables, message=FALSE}
winsize = 1000000
# Column Names
colIDs <- c("scaff","source","seqOnt","start","end","score","strand","phase",
            "attr", "count", "TElen", "eleLen", "density")
colsKeep <- c("scaff", "start", "end", "strand", "attr", "count", "TElen", 
              "eleLen", "density")

# Scaffolds (too many scaffs leads to plotting issues)
#NCfemScaffs <- c("A1", "A2", "A4", "X_NeoX")
H1scaffs <- c("A1", "A2", "A4", "X")
H2scaffs <- c("A1", "A2", "A4", "Y1", "Y2")

# Division points for the sex chromosome regions 
sexRegionEnds <- tibble(
  Region = c("PAR1", "Old", "New", "PAR2"),
  X = c(71000000, 261000000, 367000000, 483000000),
  Y1 = c(343000000, 272000000, 64000000, NA),
  Y2 = c(NA, 348000000, 159000000, 121000000)
)
```

Function to import the files with upstream, downstream, exon, and intron TE information into one list of tibbles.

```{r importFunction}
#Make a list of dataframes for each region
importFiles <- function(name) {
  # Create file list based on expected dir layout
  upstreamTEsFile <- paste("data/",name,".TEsUpstream", sep="")
  downstreamTEsFile <- paste("data/",name,".TEsDownstream", sep="")
  exonTEsFile <- paste("data/",name,".TEsExons", sep="")
  intronTEsFile <- paste("data/",name,".TEsIntrons", sep="")
  
  # Actually import the files with those names
  upstreamTEs <- read_tsv(upstreamTEsFile, col_names = colIDs) %>%
    select(all_of(colsKeep)) #%>%
    #rename(upCount = count, upDensity = density)
  downstreamTEs <- read_tsv(downstreamTEsFile, col_names = colIDs) %>%
    select(all_of(colsKeep)) #%>%
    #rename(downCount = count, downDensity = density)
  exonTEs <- read_tsv(exonTEsFile, col_names = colIDs) %>%
    select(all_of(colsKeep)) #%>%
    #rename(exonCount = count, exonDensity = density)
  intronTEs  <- read_tsv(intronTEsFile, col_names = colIDs) %>%
    select(all_of(colsKeep)) #%>% 
    #rename(intronCount = count, intronDensity = density)
  
  return( list(upstreamTEs, downstreamTEs, exonTEs, intronTEs) )
}
```

Function to window the data so it's a reasonable size for plotting (returns a tibble with total size, summed TE count, a tally of observations, summed TE length, and average TE frequency for each window).

```{r summarizeFunction}
windowAndMerge <- function(dfList, scaffList) {
  
  upTEs <- dfList[[1]] %>%
    filter(scaff %in% scaffList) %>%
    mutate(winnum = start%/%as.numeric(winsize)) %>%
    group_by(scaff, winnum) %>%
    summarise(upSize = sum(eleLen), upSum = sum(count), upNum = n(),
              upLen = sum(TElen), upAvg = mean(density))
  downTEs <- dfList[[2]] %>%
    filter(scaff %in% scaffList) %>%
    mutate(winnum = start%/%as.numeric(winsize)) %>%
    group_by(scaff, winnum) %>%
    summarise(downSize = sum(eleLen), downSum = sum(count), downNum = n(),
              downLen = sum(TElen), downAvg = mean(density))
  exTEs <- dfList[[3]] %>%
    filter(scaff %in% scaffList) %>%
    mutate(winnum = start%/%as.numeric(winsize)) %>%
    group_by(scaff, winnum) %>%
    summarise(exonSize = sum(eleLen), exonSum = sum(count), exonNum = n(),
              exonLen = sum(TElen), exonAvg = mean(density))
  inTEs <- dfList[[4]] %>%
    filter(scaff %in% scaffList) %>%
    mutate(winnum = start%/%as.numeric(winsize)) %>%
    group_by(scaff, winnum) %>%
    summarise(intronSize = sum(eleLen), intronSum = sum(count), intronNum = n(),
              intronLen = sum(TElen), intronAvg = mean(density))
    
  allTogether <- full_join(upTEs, downTEs) %>%
    full_join(exTEs) %>%
    full_join(inTEs)

  
  #returnCounts <- allTogether %>%
  #  pivot_longer(c(upSum, downSum, exonSum, intronSum),
  #               names_to = "AreaC", values_to = "sum") %>%
  #  pivot_longer(c(upAvg, downAvg, exonAvg, intronAvg),
  #               names_to = "AreaA", values_to = "avg") %>%
  #  pivot_longer(c(upNum, downNum, exonNum, intronNum),
  #               names_to = "AreaN", values_to = "count")
  
  return(allTogether)
}  
```

Merge the H1 and H2 data by gene orthology. There are usually multiple rows for introns/exons per gene - want to use the average TE density in introns/exons for a given gene.

```{r genewiseMergeFunction}
robustMin <- function(x) {if (length(x)>0) min(x) else NA}

genewise <- function(H1list, H2list, pairList) {
  {
    H1tibble <- H1list %>%
      # Flatten the list of tibbles, adding a new row to specify location
      list_rbind(, names_to = "location") %>%
      # Only use the chromosome-scale scaffolds
      filter(scaff %in% H1scaffs) %>%
      # Rename the location var so the values are meaningful & extract the
      #gene name from the attribute column to match with pair list
      mutate(location = case_when(location == 1 ~ "upstream",
                                  location == 2 ~ "downstream",
                                  location == 3 ~ "intron",
                                  location == 4 ~ "exon"),
             H1 = str_split_i( 
               str_match(attr, "^ID=(.+?);")[,2],
               pattern = "-", 1) ) %>%
      # Keep only the useful columns (don't need end, strand, or attr)
      select(location, scaff, start, count:density, H1) %>%
      # Summarise the exon/intron data per gene
      group_by(location, scaff, H1) %>%
      summarize(start = robustMin(start), count = sum(count), 
                TElen = sum(TElen), eleLen = sum(eleLen),
                frequency = mean(density), .groups = "keep") %>%
      # Find the H2 gene match for H1 orthologs (and don't keep any non-orthologs)
      inner_join(pairList, by = join_by(H1), multiple = "all")
  } # Make the H1 tibble (documented well)
  {
    H2tibble <- H2list %>%
      list_rbind(, names_to = "location") %>%
      filter(scaff %in% H2scaffs) %>%
      mutate(location = case_when(location == 1 ~ "upstream",
                                  location == 2 ~ "downstream",
                                  location == 3 ~ "intron",
                                  location == 4 ~ "exon"),
             H2 = str_split_i( 
               str_match(attr, "^ID=(.+?);")[,2],
               pattern = "-", 1) ) %>%
      select(location, scaff, start, count:density, H2) %>%
      group_by(location, scaff, H2) %>%
      summarize(start = robustMin(start), count = sum(count), 
                TElen = sum(TElen), eleLen = sum(eleLen),
                frequency = mean(density), .groups = "keep") %>%
      inner_join(pairList, by = join_by(H2), multiple = "all")
  } # Make the H2 tibble
  
  # Join the two tibbles together
  paired <- full_join(H1tibble, H2tibble, by = c("H1", "H2", "location"),
                      suffix = c("H1", "H2"), multiple = "all")
  
  return(paired) 
}
```

Functions to split up windowed data into pseudo-autosomal and old/new sex-linked regions. Very simple, but used often & I know I shouldn't copy-paste code <insert awkward guilt about past/present self>.

```{r functionsForLabellingSexRegions}
divideX <- function(df) {
  df %>%
    mutate(region = case_when(
    winnum < 71   ~ "PAR1",
    winnum < 261  ~ "Old",
    winnum < 367  ~ "New",
    winnum >= 367 ~ "PAR2"
  ))
}

divideY1 <- function(df) {
  df %>%
    mutate(region = case_when(
    winnum < 64   ~ "New",
    winnum < 272  ~ "Old",
    winnum >= 272 ~ "PAR1"
  ))
}

divideY2 <- function(df) {
  df %>%
    mutate(region = case_when(
      winnum < 121  ~ "PAR2",
      winnum < 159  ~ "New",
      winnum >= 159 ~ "Old"
    ))
}
```


## Import

Actually importing the data & summarizing it. What results is a list of tibbles (which gets summarized in the "counts" tibbles) and the list of one-to-one gene pairs.

```{r import, message=FALSE}
# NC male Haplotype 1 
H1list <- importFiles("Hap2litofftohap1") #list of tibbles with flank info
H1counts <- windowAndMerge(H1list, H1scaffs) #window the flank information

# NC male Haplotype 2
H2list <- importFiles("H2")
H2counts <- windowAndMerge(H2list, H2scaffs)

# Gene pairing information
pairList <- read_tsv("data/onetoone.pairs", col_names = c("H1","H2"))
```


## Looking at regions within the sex chromosomes

Filter the data for only the X, Y1, and Y2 information. Then summarize (sum the lengths of all TEs and divide by the total length of the region) per region linkage (pseudoautosomal, old sex-linked, or new sex-linked).

```{r labelSexRegions}
## NC haplotype 1 has the X chromosome information
XregionTEs <- H1counts %>%
  select(scaff, winnum, upLen, downLen,, exonLen, intronLen,
         upSize, downSize, exonSize, intronSize) %>%
  filter(scaff == "X") %>%
  divideX %>% #label the regions pseudo-autosomal, old sex, or new sex
  group_by(region) %>%
  summarize(up = sum(upLen, na.rm=T) / sum(upSize, na.rm=T),
            down = sum(downLen, na.rm=T) / sum(downSize, na.rm=T),
            exon = sum(exonLen, na.rm=T) / sum(exonSize, na.rm=T),
            intron = sum(intronLen, na.rm=T) / sum(intronSize, na.rm=T)) %>%
  pivot_longer(cols = up:intron, names_to = "location", values_to = "X")


##  NC haplotype 2 has the Y1 and Y2 chromosome information
Y1regionTEs <- H2counts %>%
  select(scaff, winnum, upLen, downLen,, exonLen, intronLen,
         upSize, downSize, exonSize, intronSize) %>%
  filter(scaff == "Y1") %>%
  divideY1 %>%
  group_by(region) %>%
  summarize(up = sum(upLen, na.rm=T) / sum(upSize, na.rm=T),
            down = sum(downLen, na.rm=T) / sum(downSize, na.rm=T),
            exon = sum(exonLen, na.rm=T) / sum(exonSize, na.rm=T),
            intron = sum(intronLen, na.rm=T) / sum(intronSize, na.rm=T)) %>%
  pivot_longer(cols = up:intron, names_to = "location", values_to = "Y1")

Y2regionTEs <- H2counts %>%
  select(scaff, winnum, upLen, downLen,, exonLen, intronLen,
         upSize, downSize, exonSize, intronSize) %>%
  filter(scaff == "Y2") %>%
  divideY2 %>%
  group_by(region) %>%
  summarize(up = sum(upLen, na.rm=T) / sum(upSize, na.rm=T),
            down = sum(downLen, na.rm=T) / sum(downSize, na.rm=T),
            exon = sum(exonLen, na.rm=T) / sum(exonSize, na.rm=T),
            intron = sum(intronLen, na.rm=T) / sum(intronSize, na.rm=T)) %>%
  pivot_longer(cols = up:intron, names_to = "location", values_to = "Y2")


## Combine the separate sex chromosome information
allRegionTEs <- full_join(XregionTEs, Y1regionTEs, 
                          by = join_by(region, location)) %>%
  full_join(Y2regionTEs, by = join_by(region, location)) %>%
  pivot_longer(cols = c(X, Y1, Y2), 
               names_to = "Chromosome", values_to = "Frequency")

# Clean up the intermediate files
rm(XregionTEs, Y1regionTEs, Y2regionTEs)
```

This is mostly to validate the data - I'll ditch it later!

```{r sexRegionSizes}
XregionSizes <- H1counts %>%
  select(scaff, winnum, upSize, downSize, exonSize, intronSize) %>%
  filter(scaff == "X") %>%
  divideX %>%
  group_by(region) %>%
   summarize(up = sum(upSize, na.rm=T), 
             down = sum(downSize, na.rm=T),
             exon = sum(exonSize, na.rm=T), 
             intron = sum(intronSize, na.rm=T)) %>%
  pivot_longer(cols = up:intron, names_to = "location", values_to = "X")


Y1regionSizes <- H2counts %>%
  select(scaff, winnum, upLen, downLen,, exonLen, intronLen,
         upSize, downSize, exonSize, intronSize) %>%
  filter(scaff == "Y1") %>%
  divideY1 %>%
  group_by(region) %>%
  summarize(up = sum(upSize, na.rm=T),
            down = sum(downSize, na.rm=T),
            exon = sum(exonSize, na.rm=T),
            intron = sum(intronSize, na.rm=T)) %>%
  pivot_longer(cols = up:intron, names_to = "location", values_to = "Y1")

Y2regionSizes <- H2counts %>%
  select(scaff, winnum, upLen, downLen,, exonLen, intronLen,
         upSize, downSize, exonSize, intronSize) %>%
  filter(scaff == "Y2") %>%
  divideY2 %>%
  group_by(region) %>%
  summarize(up = sum(upSize, na.rm=T),
            down = sum(downSize, na.rm=T),
            exon = sum(exonSize, na.rm=T),
            intron = sum(intronSize, na.rm=T)) %>%
  pivot_longer(cols = up:intron, names_to = "location", values_to = "Y2")


## Combine size information
allRegionSizes <- full_join(XregionSizes, Y1regionSizes, 
                          by = join_by(region, location)) %>%
  full_join(Y2regionSizes, by = join_by(region, location)) %>%
  pivot_longer(cols = c(X, Y1, Y2), 
               names_to = "Chromosome", values_to = "Size")

# Clean up intermediate files
rm(XregionSizes, Y1regionSizes, Y2regionSizes)
```



## Per-Gene Comparisons

Want to look at differences per gene instead of in aggregate!

```{r geneByGene, warning=FALSE}
pairedGenes <- genewise(H1list, H2list, pairList)

# Trying to figure out why NAs appear during the join - the input was supposed to include *only* H1:H2 orthologs...
# paired %>%
#   filter(is.na(countH1)) #396 rows
# paired %>%
#   filter(is.na(countH2)) #649 rows
# 
# H1tibble %>%
#   filter(is.na(H2)) #0
# H2tibble %>%
#   filter(is.na(H1)) #0
```

If we just want to look at the genes with orthologs on X or Y1/Y2



# Visualization

## Across Chromosomes {.tabset}

### Averaged TE Frequency

Bedtools provides a TE *frequency* measurement (how much of the feature's length overlaps with an annotated TE), which I averaged within each 1Mb window.

```{r averageTEfrequency}
H1counts %>%
  ggplot(aes(winnum)) +
  geom_line(aes(y = upAvg, colour = "Upstream")) +
  geom_line(aes(y = downAvg, colour = "Downstream")) +
  geom_line(aes(y = exonAvg, colour = "Exon")) +
  geom_line(aes(y = intronAvg, colour = "Intron")) +
  pubTheme + theme(legend.position=c(0.87,0.5)) +
  labs(x = "H1 Position (1Mb windows)", 
       y = "Average TE Frequency in this Window",
       colour = "Location") +
  facet_grid( vars(scaff), scale = "free_x", space = "free" )

H2counts %>%
  ggplot(aes(winnum)) +
  geom_line(aes(y = upAvg, colour = "Upstream")) +
  geom_line(aes(y = downAvg, colour = "Downstream")) +
  geom_line(aes(y = exonAvg, colour = "Exon")) +
  geom_line(aes(y = intronAvg, colour = "Intron")) +
  pubTheme + theme(legend.position=c(0.87,0.5)) +
  labs(x = "H2 Position (1Mb windows)", 
       y = "Average TE Frequency in this Window",
       colour = "Location") +
  facet_grid( vars(scaff), scale = "free_x", space = "free" )
```

Generally, the 1kb regions up/downstream of genes contain a larger proportion of TEs than introns or exons do. There are some interestingly TE-ridden introns, though, too.

### TE Frequency (re-calculated)

Instead of using the mean TE frequency for each window, re-calculate the frequency by dividing the summed TE length for each window by the summed length of all elements (introns, upstream/downstream regions) for each window.

```{r notAverageTEfrequency}
H1counts %>%
  ggplot(aes(winnum)) +
  geom_line(aes(y = upLen / upSize, colour = "Upstream")) +
  geom_line(aes(y = downLen / downSize, colour = "Downstream")) +
  geom_line(aes(y = exonLen / exonSize, colour = "Exon")) +
  geom_line(aes(y = intronLen / intronSize, colour = "Intron")) +
  pubTheme + theme(legend.position=c(0.87,0.5)) +
  labs(x = "H1 Position (1Mb windows)", y = "TE Frequency",
       colour = "Location") +
  facet_grid( vars(scaff), scale = "free_x", space = "free" )

H2counts %>%
  ggplot(aes(winnum)) +
  geom_line(aes(y = upLen / upSize, colour = "Upstream")) +
  geom_line(aes(y = downLen / downSize, colour = "Downstream")) +
  geom_line(aes(y = exonLen / exonSize, colour = "Exon")) +
  geom_line(aes(y = intronLen / intronSize, colour = "Intron")) +
  pubTheme + theme(legend.position=c(0.87,0.5)) +
  labs(x = "H2 Position (1Mb windows)", y = "TE Frequency",
       colour = "Location") +
  facet_grid( vars(scaff), scale = "free_x", space = "free" )
```

### TE Count

Bedtools overlap also provides the *number* of TEs in a given feature, which is slightly more relevant to my interest in recent insertions. I averaged this sum within each 1Mb window. 

```{r plotTEsum}
H1counts %>%
  ggplot(aes(winnum)) +
  geom_line(aes(y = upSum, colour = "Upstream")) +
  geom_line(aes(y = downSum, colour = "Downstream")) +
  geom_line(aes(y = exonSum, colour = "Exon")) +
  geom_line(aes(y = intronSum, colour = "Intron")) +
  pubTheme + theme(legend.position=c(0.87,0.5)) +
  labs(x = "H1 Position (1Mb windows)", 
       y = "Windowed TE count",
       colour = "Location") +
  facet_grid( vars(scaff), scale = "free_x", space = "free" )

H2counts %>%
  ggplot(aes(winnum)) +
  geom_line(aes(y = upSum, colour = "Upstream")) +
  geom_line(aes(y = downSum, colour = "Downstream")) +
  geom_line(aes(y = exonSum, colour = "Exon")) +
  geom_line(aes(y = intronSum, colour = "Intron")) +
  pubTheme + theme(legend.position=c(0.87,0.5)) +
  labs(x = "H2 Position (1Mb windows)", 
       y = "Windowed TE count",
       colour = "Location") +
  facet_grid( vars(scaff), scale = "free_x", space = "free" )
```

Occasional huge spikes in the number of TEs in H1 introns made things hard to interpret for the rest of the features.
 
```{r plotH1sumWithoutIntrons}
H1counts %>%
  ggplot(aes(winnum)) +
  geom_line(aes(y = upSum, colour = "Upstream")) +
  geom_line(aes(y = downSum, colour = "Downstream")) +
  geom_line(aes(y = exonSum, colour = "Exon")) +
#  geom_line(aes(y = intronSum, colour = "Intron")) +
  pubTheme + theme(legend.position=c(0.87,0.5)) +
  labs(x = "H1 Position (1Mb windows)", 
       y = "Windowed TE count",
       colour = "Location") +
  facet_grid( vars(scaff), scale = "free_x", space = "free" )
```

I also divided each window's TE sum metric by the number of features in it (e.g. the average number of intron TEs in the window is divided by the number of introns).

```{r plotCountOverFeatureCount}
H1counts %>%
  ggplot(aes(winnum)) +
  geom_line(aes(y = upSum/upNum, colour = "Upstream")) +
  geom_line(aes(y = downSum/downNum, colour = "Downstream")) +
  geom_line(aes(y = exonSum/exonNum, colour = "Exon")) +
  geom_line(aes(y = intronSum/intronNum, colour = "Intron")) +
  pubTheme + theme(legend.position=c(0.87,0.5)) +
  labs(x = "H1 Position (1Mb windows)", 
       y = "Windowed count divided by feature count",
       colour = "Location") +
  facet_grid( vars(scaff), scale = "free_x", space = "free" )

H2counts %>%
  ggplot(aes(winnum)) +
  geom_line(aes(y = upSum/upNum, colour = "Upstream")) +
  geom_line(aes(y = downSum/downNum, colour = "Downstream")) +
  geom_line(aes(y = exonSum/exonNum, colour = "Exon")) +
  geom_line(aes(y = intronSum/intronNum, colour = "Intron")) +
  pubTheme + theme(legend.position=c(0.87,0.5)) +
  labs(x = "H2 Position (1Mb windows)", 
       y = "Windowed count divided by feature count",
       colour = "Location") +
  facet_grid( vars(scaff), scale = "free_x", space = "free" )



countOverFeatureCount <- H1counts %>%
  filter(scaff == "X") %>%
  ggplot(aes(winnum)) +
  geom_line(aes(y = upSum/upNum, colour = "Upstream")) +
  geom_line(aes(y = downSum/downNum, colour = "Downstream")) +
  geom_line(aes(y = exonSum/exonNum, colour = "Exon")) +
  geom_line(aes(y = intronSum/intronNum, colour = "Intron")) +
  geom_line(data = H2counts, aes(y = upSum/upNum, colour = "Upstream")) +
  geom_line(data = H2counts, aes(y = downSum/downNum, colour = "Downstream")) +
  geom_line(data = H2counts, aes(y = exonSum/exonNum, colour = "Exon")) +
  geom_line(data = H2counts, aes(y = intronSum/intronNum, colour = "Intron")) +
  pubTheme + #theme(legend.direction = "horizontal", legend.position = "top") +
  labs(x = "Position (1Mb windows)", 
       y = "Windowed count divided by feature count",
       colour = "Location") +
  facet_grid(vars(scaff), scale="free_x", space="free")

countOverFeatureCount
```

With the unfiltered data set, there were an average of 2 TEs within 1kb of genes. With the filtered set, it looks about the same but with variation closer to what I'd expect (low levels in the regions of high gene density).


### Introns are Weird

I thought this section was going to be unhelpful with the filtered data set, but this is definitely helping to look into what's going on with the introns on the X chromosome.

```{r INTRONScomparingNumAndSumFor}
H1counts %>%
  select(c(scaff, winnum, intronNum, intronSum)) %>%
  ggplot(aes(x = winnum)) +
  geom_line(aes(y = intronNum, colour = "Introns")) +
  geom_line(aes(y = intronSum, colour = "TEs in Introns")) +
  pubTheme + theme(legend.position=c(0.87,0.5)) +
  labs(x = "H1 Position (1Mb windows)", y = "Summed count in window",
       colour = "Feature Type") +
  facet_grid( vars(scaff), scale = "free_x", space = "free" )

H2counts %>%
  select(c(scaff, winnum, intronNum, intronSum)) %>%
  ggplot(aes(x = winnum)) +
  geom_line(aes(y = intronNum, colour = "Introns")) +
  geom_line(aes(y = intronSum, colour = "TEs in Introns")) +
  pubTheme + theme(legend.position=c(0.87,0.5)) + ylim(0,500) +
  labs(x = "H2 Position (1Mb windows)", y = "Summed count in Window",
       colour = "Feature Type") +
  facet_grid( vars(scaff), scale = "free_x", space = "free" )
```

The H1 intron data fits really nicely with my expectations but I find H2 intron data very confusing. I don't know why there are so many fewer introns in the H2 gene annotation...


```{r INTRONScomparingLengths}
H1counts %>%
  select(c(scaff, winnum, intronSize, intronLen)) %>%
  ggplot(aes(x = winnum)) +
  geom_line(aes(y = intronSize, colour = "Introns")) +
  geom_line(aes(y = intronLen, colour = "TEs in Introns")) +
  pubTheme + theme(legend.position=c(0.87,0.5)) +
  labs(x = "H1 Position (1Mb windows)", y = "Summed length in window",
       colour = "Feature Type") +
  facet_grid( vars(scaff), scale = "free_x", space = "free" )

H2counts %>%
  select(c(scaff, winnum, intronSize, intronLen)) %>%
  ggplot(aes(x = winnum)) +
  geom_line(aes(y = intronSize, colour = "Introns")) +
  geom_line(aes(y = intronLen, colour = "TEs in Introns")) +
  pubTheme + theme(legend.position=c(0.87,0.5)) +
  labs(x = "H2 Position (1Mb windows)", y = "Summed length in Window",
       colour = "Feature Type") +
  facet_grid( vars(scaff), scale = "free_x", space = "free" )
```


## Bar plots {.tabset}

### Region Sizes

```{r barplotSexChromSizes, warning=FALSE}
regionSizes <- ggplot(allRegionSizes, aes(x = region, y = Size, fill = Chromosome)) +
  geom_col(position = position_dodge()) +
  pubTheme + scale_fill_manual(values = pubColours) +
  labs(x = "Region", y = "Number of bp in this region", 
       fill = "Chromosome") +
  facet_grid(vars(location), space = "free")

regionSizes
```

X chromosome introns seem to be much more frequent/larger for these 1:1 orthologs. 

### Divisions within sex chromosomes

```{r barplotsSexChroms, warning=FALSE}
regionFrequency <- allRegionTEs %>%
  ggplot(aes(x = region, y = Frequency, fill = Chromosome)) +
  geom_col(position = position_dodge()) +
  pubTheme + scale_fill_manual(values = pubColours) +
  labs(x = "Region", y = "Frequency of TEs", 
       fill = "Chromosome") +
  facet_grid(vars(location), space = "free")

regionFrequency
```

The number of introns in genes on the X is very high, which *might* help explain why there are so many TEs in them?? Overall, this is just super weird...

## Genewise Comparisons {.tabset}

For this section, every point on the graph is a gene with at least one ortholog in Haplotype 1 and Haplotype 2.

### Quick stats - Count

```{r genewiseQuickCountStats}
#glimpse(pairedGenes)

# Count data
pairedCount <- pairedGenes %>%
  group_by(scaffH1, scaffH2, location) %>%
  select(location:scaffH2, countH1, countH2) %>%
  summarise(meanH1 = mean(countH1), meanH2 = mean(countH2), n = n(),
           medianH1 = median(countH1), medianH2 = median(countH2),
           .groups = "keep") %>%
  drop_na() # still very concerned about this

#glimpse(pairedCount)


pairedGenes %>%
  group_by(scaffH1, scaffH2, location) %>%
  select(location:scaffH2, countH1, countH2) %>%
  drop_na() %>%
  pivot_longer(cols = c(countH1, countH2), names_to = "Haplotype", 
               values_to = "TEcount") %>%
  mutate(Haplotype = case_when(Haplotype == "countH1" ~ "H1",
                               Haplotype == "countH2" ~ "H2")) %>%
  ggplot() +
  geom_violin(aes(x = scaffH2, y = TEcount, fill = Haplotype)) +
  #geom_point(aes(x = scaffH2, y = TEcount, colour = Haplotype), alpha = 0.1,
  #           position = position_dodge2(width = 0.5)) +
  pubTheme + theme(legend.position = c(0.98, 0.89)) +
  scale_colour_manual(values = pubColours) +
  scale_fill_manual(values = pubColours) + 
  labs(x = "Chromosome", y = "TE count") +
  facet_wrap(vars(location), scales = "free")
```

```{r genewisePlotQuickCountStats}
# Plot Mean 
ggplot(pairedCount) +
  geom_point(aes(x = scaffH1, y = meanH1, size = n), colour = pubColours[1],
             position = position_dodge2(width = 0.3)) +
  geom_point(aes(x = scaffH2, y = meanH2, size = n), colour = pubColours[2],
             position = position_dodge2(width = 0.3)) +
  # Adding text to know where the ortholog is
  geom_text(aes(x = scaffH1, y = meanH1, label = scaffH2), colour = pubColours[5],
             position = position_dodge2(width = 0.3), size = 3) +
  geom_text(aes(x = scaffH2, y = meanH2, label = scaffH1), colour = pubColours[5],
             position = position_dodge2(width = 0.3), size = 3) +
  pubTheme + theme(legend.position = "none") +
  labs(title = "Mean TE count",x = "Scaffold", y = "Mean TE count") +
  facet_wrap(vars(location), scales = "free")

# Plot median
ggplot(pairedCount) +
  geom_point(aes(x = scaffH1, y = medianH1, size = n), colour = pubColours[1],
             position = position_dodge2(width = 0.3)) +
  geom_point(aes(x = scaffH2, y = medianH2, size = n), colour = pubColours[2],
             position = position_dodge2(width = 0.3)) +
  # Adding text to know where the ortholog is
  geom_text(aes(x = scaffH1, y = medianH1, label = scaffH2), colour = pubColours[5],
             position = position_dodge2(width = 0.3), size = 3) +
  geom_text(aes(x = scaffH2, y = medianH2, label = scaffH1), colour = pubColours[5],
             position = position_dodge2(width = 0.3), size = 3) +
  pubTheme + theme(legend.position = "none") +
  labs(title = "Median TE count",x = "Scaffold", y = "Median TE count") +
  facet_wrap(vars(location), scales = "free")
```

### Quick Stats - Frequency

Count feels like a more reasonable metric to do this with, but we're using frequency for everything else.

```{r genewiseQuickFrequencyStats}
pairedFreq <- pairedGenes %>%
  group_by(scaffH1, scaffH2, location) %>%
  select(location:scaffH2, frequencyH1, frequencyH2) %>%
  summarise(meanH1 = mean(frequencyH1), meanH2 = mean(frequencyH2), n = n(),
            medianH1 = median(frequencyH1), medianH2 = median(frequencyH2),
            .groups = "keep") %>%
  drop_na() # still very concerned about this

glimpse(pairedFreq)

pairedGenes %>%
  group_by(scaffH1, scaffH2, location) %>%
  select(location:scaffH2, frequencyH1, frequencyH2) %>%
  drop_na() %>%
  pivot_longer(cols = c(frequencyH1, frequencyH2), names_to = "Haplotype", 
               values_to = "frequency") %>%
  mutate(Haplotype = case_when(Haplotype == "densityH1" ~ "H1",
                               Haplotype == "densityH2" ~ "H2")) %>%
  ggplot() +
  geom_boxplot(aes(x = scaffH2, y = frequency, fill = Haplotype)) +
  geom_point(aes(x = scaffH2, y = frequency, colour = Haplotype), 
             alpha = 0.02, position = "jitter") +
  pubTheme + theme(legend.position = c(0.9, 0.85)) +
  scale_colour_manual(values = pubColours) +
  scale_fill_manual(values = pubColours) + 
  labs(x = "H2 Chromosome", y = "TE frequency") +
  facet_wrap(vars(location), scales = "free")
```


```{r genewisePlotQuickFreqStats}
# Plot Mean 
ggplot(pairedFreq) +
  geom_point(aes(x = scaffH1, y = meanH1, size = n), colour = pubColours[1],
             position = position_dodge2(width = 0.3)) +
  geom_point(aes(x = scaffH2, y = meanH2, size = n), colour = pubColours[2],
             position = position_dodge2(width = 0.3)) +
  # Adding text to know where the ortholog is
  geom_text(aes(x = scaffH1, y = meanH1, label = scaffH2), colour = pubColours[5],
             position = position_dodge2(width = 0.3), size = 3) +
  geom_text(aes(x = scaffH2, y = meanH2, label = scaffH1), colour = pubColours[5],
             position = position_dodge2(width = 0.3), size = 3) +
  pubTheme + #theme(legend.position = "none") +
  labs(x = "Scaffold", y = "Mean TE frequency") +
  facet_wrap(vars(location), scales = "free")

# Plot median
ggplot(pairedFreq) +
  geom_point(aes(x = scaffH1, y = medianH1, size = n), colour = pubColours[1],
             position = position_dodge2(width = 0.3)) +
  geom_point(aes(x = scaffH2, y = medianH2, size = n), colour = pubColours[2],
             position = position_dodge2(width = 0.3)) +
  # Adding text to know where the ortholog is
  geom_text(aes(x = scaffH1, y = medianH1, label = scaffH2), colour = pubColours[5],
             position = position_dodge2(width = 0.3), size = 3) +
  geom_text(aes(x = scaffH2, y = medianH2, label = scaffH1), colour = pubColours[5],
             position = position_dodge2(width = 0.3), size = 3) +
  pubTheme + #theme(legend.position = "none") +
  labs(x = "Scaffold", y = "Median TE frequency") +
  facet_wrap(vars(location), scales = "free")
```

### Orthologs on sex chromosomes

Look for a difference in TE frequency near genes with orthologs on the X and Y1/Y2.

```{r genewiseSexChromosomesPoints, message=FALSE}
pairedGenes %>%
  filter(scaffH2 == "Y1" | scaffH2 == "Y2") %>%
  filter(scaffH1 == "X") %>%
  ggplot() +
  geom_point(aes(x = frequencyH1, y = frequencyH2, colour = scaffH2), 
             alpha=0.3) +
  geom_smooth(aes(x = frequencyH1, y = frequencyH2, colour = scaffH2)) +
  geom_abline() +
  pubTheme + scale_colour_manual(values = pubColours) + 
  #theme(legend.position = "none") +
  labs(title = "Genes with a Y1/Y2 ortholog in Hap2 and an X ortholog in Hap1",
       x = "TE frequency for H1 ortholog", y = "TE frequency for H2 ortholog", 
       colour = "H2 Chrom") +
  facet_wrap(vars(location), nrow = 2)



# There are no longer genes that appear on the Y chromosomes but not on the X

pairedGenes %>%
  filter(scaffH2 == "Y1" | scaffH2 == "Y2") %>%
  filter(scaffH1 != "X") #%>%
#   ggplot() +
#   geom_point(aes(x = frequencyH1, y = frequencyH2, colour = scaffH1), 
#              alpha=0.3) +
#   geom_smooth(aes(x = frequencyH1, y = frequencyH2, colour = scaffH1)) +
#   geom_abline() +
#   pubTheme + scale_colour_manual(values = pubColours) + 
#   #theme(legend.position = "none") +
#   labs(title = "Genes with a Y1/Y2 ortholog in Hap2 and autosomal ortholog in Hap1",
#        x = "TE frequency for H1 ortholog", y = "TE frequency for H2 ortholog", 
#        colour = "H1 Chrom") +
#   #ylim(0,1) +
#   facet_wrap(vars(location))
```

The results feel very confusing to me

```{r barplotGenewise, message=FALSE}
pairedGenes %>%
  filter(scaffH2 == "Y1" | scaffH2 == "Y2") %>%
  filter(scaffH1 == "X") %>%
  mutate(frequencyRatio = frequencyH1/frequencyH2) %>%
  ungroup() %>%
  select(scaffH1, scaffH2, frequencyRatio, location) %>%
  #pivot_longer(cols = c(scaffH1, scaffH2), names_to = "scaff") #%>%
  ggplot() +
  geom_point(aes(x = location, y = frequencyRatio)) +
  #geom_col(aes(x = location, y = frequencyRatio, fill = scaffH2), 
  #         position = position_dodge()) +
  pubTheme + scale_colour_manual(values = pubColours) + 
  #theme(legend.position = "none") +
  labs(title = "Genes with a Y1/Y2 ortholog in Hap2 and an X ortholog in Hap1",
       x = "TE frequency for H1 ortholog", y = "Ratio of H1 TEs / H2 TEs", 
       colour = "H2 Chrom") #+
  facet_wrap(vars(location), nrow = 2)
```



### Chromosomal Landscape

```{r filterCompressForPlotting}
pairedTest <- pairedGenes %>%
  filter(scaffH1 == scaffH2 | 
           scaffH1 == "X" | scaffH2 == "Y1" | scaffH2 == "Y2") %>%
  mutate(winnum = startH1 %/% as.numeric(winsize)) %>%
  group_by(location, scaffH1, winnum) %>%
  summarise(freqH1 = mean(frequencyH1), freqH2 = mean(frequencyH2),
            .groups = "keep") %>% 
  drop_na(freqH1)
```


Green is H2 and blue is H1

```{r genewiseLandscapeH1H2}
pairedTest %>%
  filter(location == "upstream") %>%
  ggplot() +
  geom_line(aes(x = winnum, y = freqH2), colour = pubColours[2]) +
  geom_line(aes(x = winnum, y = freqH1), colour = pubColours[1]) +
  scale_colour_manual(values = pubColours) +
  pubTheme + theme(legend.position = c(0.85, 0.5)) +
  labs(x = "Position (1Mb windows)", y = "Upstream TE frequency") +
  facet_grid(vars(scaffH1), scales = "free")

pairedTest %>%
  filter(location == "downstream") %>%
  ggplot() +
  geom_line(aes(x = winnum, y = freqH2), colour = pubColours[2]) +
  geom_line(aes(x = winnum, y = freqH1), colour = pubColours[1]) +
  scale_colour_manual(values = pubColours) +
  pubTheme + theme(legend.position = c(0.85, 0.5)) +
  labs(x = "Position (1Mb windows)", y = "Downstream TE frequency") +
  facet_grid(vars(scaffH1), scales = "free")

pairedTest %>%
  filter(location == "exon") %>%
  ggplot() +
  geom_line(aes(x = winnum, y = freqH2), colour = pubColours[2]) +
  geom_line(aes(x = winnum, y = freqH1), colour = pubColours[1]) +
  scale_colour_manual(values = pubColours) +
  pubTheme + theme(legend.position = c(0.85, 0.5)) +
  labs(x = "Position (1Mb windows)", y = "Exon TE frequency") +
  facet_grid(vars(scaffH1), scales = "free")

pairedTest %>%
  filter(location == "intron") %>%
  ggplot() +
  geom_line(aes(x = winnum, y = freqH2), colour = pubColours[2]) +
  geom_line(aes(x = winnum, y = freqH1), colour = pubColours[1]) +
  scale_colour_manual(values = pubColours) +
  pubTheme + theme(legend.position = c(0.85, 0.5)) +
  labs(x = "Position (1Mb windows)", y = "Intron TE frequency") +
  facet_grid(vars(scaffH1), scales = "free")
```

```{r}
ggplot(pairedTest) +
  geom_line(aes(x = winnum, y = freqH1, colour = location)) +
  scale_colour_manual(values = pubColours) +
  pubTheme + theme(legend.position = c(0.85, 0.5)) +
  labs(x = "Position (1Mb windows)", y = "H1 TE frequency",
       colour = "Location") +
  facet_grid(vars(scaffH1), scales = "free")

ggplot(pairedTest) +
  geom_line(aes(x = winnum, y = freqH2, colour = location)) +
  scale_colour_manual(values = pubColours) +
  pubTheme + theme(legend.position = c(0.85, 0.5)) +
  labs(x = "Position (1Mb windows)", y = "H2 TE frequency", 
       colour = "Location") +
  facet_grid(vars(scaffH1), scales = "free")
```




# Figure 4

```{r fig4}
fig4 <- 
  countOverFeatureCount / (regionFrequency | regionSizes) +
  plot_layout(heights = c(2, 1)) +
  plot_layout(guides = "collect") +
  plot_annotation(
  title = "Figure 4",
  tag_levels = "A", 
  caption = "(A) The fractio (B) Fraction of the region occupied by Genes.  (C) Fraction of the region occupied by TEs.")

fig4 #quick preview to double-check

# Save the plot as figure3.png "with sensible defaults"
ggsave("figure4.png", plot = fig4, width = 9, height = 7)
```



# Unused

Testing for genewise function 

```{r testingGenewise, eval=FALSE}
#rm(genePairs, inPairs, upPairs, allPairs, pair1, pair2, pairList)

{
  gene2 <- pairList$H2[1]
#pattern2 <- paste("^ID=",gene2, sep = "")

pair1 <- H1list[[3]] %>%
  filter( str_detect(attr, gene1) ) %>%
  #distinct() %>%
  select(scaff, start, count:density) %>%
  mutate("geneH1" = gene1, "geneH2" = gene2) %>%
  group_by(scaff, geneH1, geneH2) %>%
  summarize(start = min(start), count = sum(count), TElen = sum(TElen), 
            eleLen = sum(eleLen),density = mean(density))
pair2 <- H2list[[3]] %>%
  filter( str_detect(attr, gene2) ) %>%
  #distinct() %>%
  select(!end:attr) %>%
  mutate("gene1" = gene1, "gene2" = gene2) 

full_join(pair1, pair2, by = c("gene1", "gene2"))
} #small test sets

{
    for (i in 1:len) {
    # Save the pair's gene name for H1 and H2
    gene1 <- pairList$H1[i]
    gene2 <- pairList$H2[i]
    #print(gene1)
    
    # Each gene will only have one upstream region and one downstream region
    for (j in 1:2) {
      pair1 <- H1list[[j]] %>%
        filter( str_detect(attr, gene1) ) %>%
        distinct() %>%
        select(scaff, start, count:density) %>%
        mutate("geneH1" = gene1, "geneH2" = gene2, "location" = j)
      print(pair1)
      pair2 <- H2list[[j]] %>%
        filter( str_detect(attr, gene2) ) %>%
        distinct() %>%
        select(scaff, start, count:density) %>%
        mutate("geneH2" = gene2, "geneH1" = gene1, "location" = j)
      print(pair2)
      allPairs[[i]] <- full_join(pair1, pair2,
                                      by = c("geneH1", "geneH2", "location"),
                                      suffix = c("H1", "H2"))
    } #end upstream/downstream search (allPairs[[1]] and allPairs[[2]] done)
    
    # There may be several exons and introns per gene - sum/average the values
    for (j in 3:4) {
      pair1 <- H1list[[j]] %>%
        filter( str_detect(attr, gene1) ) %>%
        distinct() %>%
        select(scaff, start, count:density) %>%
        mutate("geneH1" = gene1, "geneH2" = gene2, "location" = j)  %>%
        group_by(scaff, geneH1, geneH2, location) %>%
        summarize(start = robustMin(start), count = sum(count), 
                  TElen = sum(TElen), eleLen = sum(eleLen),
                  density = mean(density), .groups = "keep")
      print(pair1)
      pair2 <- H2list[[j]] %>%
        filter( str_detect(attr, gene2) ) %>%
        distinct() %>%
        select(scaff, start, count:density) %>%
        mutate("geneH2" = gene2, "geneH1" = gene1, "location" = j) %>%
        group_by(scaff, geneH1, geneH2, location) %>%
        summarize(start = robustMin(start), count = sum(count), 
                  TElen = sum(TElen), eleLen = sum(eleLen),
                  density = mean(density), .groups = "keep")
      print(pair2)
      allPairs[[j]][[i]] <- full_join(pair1, pair2, 
                                      by = c("geneH1", "geneH2", "location"),
                                      suffix = c("H1", "H2"))
    } #end exon/intron search
    
  } #end gene pair search
} #iterating over H1list and H2list

{
upPair1 <- H1list[[1]] %>%
      filter( str_detect(attr, gene1) ) %>%
      select(scaff, count:density) %>%
      mutate("gene1" = gene1, "gene2" = gene2,
             "count1" = count, "TElen1" = TElen, 
             "eleLen1" = eleLen, "density1" = density) %>%
      select(!count:density)
    downPair1 <- H1list[[2]] %>%
      filter( str_detect(attr, gene1) ) %>%
      select(scaff, count:density) %>%
      mutate("gene1" = gene1, "gene2" = gene2,
             "count1" = count, "TElen1" = TElen, 
             "eleLen1" = eleLen, "density1" = density) %>%
      select(!count:density)
    exPair1 <- H1list[[3]] %>%
      filter( str_detect(attr, gene1) ) %>%
      select(scaff, count:density) %>%
      mutate("gene1" = gene1, "gene2" = gene2,
             "count1" = count, "TElen1" = TElen, 
             "eleLen1" = eleLen, "density1" = density) %>%
      select(!count:density)
    inPair1 <- H1list[[4]] %>%
      filter( str_detect(attr, gene1) ) %>%
      select(scaff, count:density) %>%
      mutate("gene1" = gene1, "gene2" = gene2,
             "count1" = count, "TElen1" = TElen, 
             "eleLen1" = eleLen, "density1" = density) %>%
      select(!count:density)
      
    upPair2 <- H2list[[1]] %>%
      filter( str_detect(attr, gene2) ) %>%
      select(scaff, count:density) %>%
      mutate("gene1" = gene1, "gene2" = gene2,
             "count2" = count, "TElen2" = TElen, 
             "eleLen2" = eleLen, "density2" = density) %>%
      select(!count:density)
    downPair2 <- H2list[[2]] %>%
      filter( str_detect(attr, gene2) ) %>%
      select(scaff, count:density) %>%
      mutate("geneH1" = gene1, "geneH2" = gene2,
             "countH2" = count, "TElenH2" = TElen, 
             "eleLenH2" = eleLen, "densityH2" = density) %>%
      select(!count:density)
    exPair2 <- H2list[[3]] %>%
      filter( str_detect(attr, gene2) ) %>%
      select(scaff, count:density) %>%
      mutate("gene1" = gene1, "gene2" = gene2,
             "count2" = count, "TElen2" = TElen, 
             "eleLen2" = eleLen, "density2" = density) %>%
      select(!count:density)
    inPair2 <- H2list[[4]] %>%
      filter( str_detect(attr, gene2) ) %>%
      select(scaff, count:density) %>%
      mutate("gene1" = gene1, "gene2" = gene2,
             "count2" = count, "TElen2" = TElen, 
             "eleLen2" = eleLen, "density2" = density) %>%
      select(!count:density)
    
        
    upPairs <- full_join(upPair1, upPair2, by = c("scaff", "gene1", "gene2"))
    downPairs <- full_join(upPair1, upPair2, by = c("scaff", "gene1", "gene2"))
    exPairs <- full_join(exPair1, exPair2, by = c("scaff", "gene1", "gene2"))
    inPairs <- full_join(inPair1, inPair2, by = c("scaff", "gene1", "gene2"))
  } # too many things
```



Different approach to summing TE bp in different regions

```{r summingBPforBarplots, eval=FALSE}
#%>%
  group_by(seqOnt, region) %>%
  summarize(summedbp = sum(bp)) %>%
  mutate(X = case_when(
    region == "PAR1"   ~ summedbp/71000000,
    region == "Old"    ~ summedbp/(261000000-71000000),
    region == "New"    ~ summedbp/(367000000-261000000),
    region == "PAR2"   ~ summedbp/(483000000-367000000)
  ))


%>%
  group_by(seqOnt, region) %>%
  summarize(summedbp = sum(bp)) %>%
  mutate(Y1 = case_when(
    region == "New"   ~ summedbp/64000000,
    region == "Old"    ~ summedbp/(272000000-64000000),
    region == "PAR1"    ~ summedbp/(343000000-272000000)
  ))


%>%
  group_by(seqOnt, region) %>%
  summarize(summedbp = sum(bp)) %>%
  mutate(Y2 = case_when(
    region == "PAR2"   ~ summedbp/121000000,
    region == "New"    ~ summedbp/(159000000-121000000),
    region == "Old"    ~ summedbp/(348000000-159000000)
  ))


allSummedTEs <- full_join(XsummedTEs, Y1summedTEs, 
                          by = join_by(seqOnt, region)) %>%
  full_join(Y2summedTEs, by = join_by(seqOnt, region)) %>%
  pivot_longer(cols = c(X, Y1, Y2), 
               names_to = "Chromosome", values_to = "bp")
```


NC female data

```{r NCfem, eval=FALSE}
# NC female
#NCfemList <- importFiles("NCfem")
#femCounts <- windowAndMerge(NCfemList, NCfemScaffs)





#femCounts %>%
#  ggplot(aes(winnum)) +
#  geom_line(aes(y = upAvg, colour = "Upstream")) +
#  geom_line(aes(y = downAvg, colour = "Downstream")) +
#  geom_line(aes(y = exonAvg, colour = "Exon")) +
#  geom_line(aes(y = intronAvg, colour = "Intron")) +
#  pubTheme + theme(legend.position=c(0.87,0.5)) +
#  labs(x = "NCfem Position (1Mb windows)", 
#       y = "Average TE Frequency in this Window",
#       colour = "Location") +
#  facet_grid( vars(scaff), scale = "free_x", space = "free" )


#femCounts %>%
#  ggplot(aes(winnum)) +
#  geom_line(aes(y = upSum, colour = "Upstream")) +
#  geom_line(aes(y = downSum, colour = "Downstream")) +
#  geom_line(aes(y = exonSum, colour = "Exon")) +
#  geom_line(aes(y = intronSum, colour = "Intron")) +
#  pubTheme + theme(legend.position=c(0.87,0.5)) +
#  labs(x = "NCfem Position (1Mb windows)", 
#       y = "Windowed TE count",
#       colour = "Location") +
#  facet_grid( vars(scaff), scale = "free_x", space = "free" )

# limited Y axis
#femCounts %>%
#  ggplot(aes(winnum)) +
#  geom_line(aes(y = upSum, colour = "Upstream")) +
#  geom_line(aes(y = downSum, colour = "Downstream")) +
#  geom_line(aes(y = exonSum, colour = "Exon")) +
#  geom_line(aes(y = intronSum, colour = "Intron")) +
#  pubTheme + theme(legend.position=c(0.87,0.5)) +
#  labs(x = "NCfem Position (1Mb windows)", 
#       y = "Windowed TE count",
#       colour = "Location") +
#  facet_grid( vars(scaff), scale = "free_x", space = "free" )

femCounts %>%
  ggplot(aes(winnum)) +
  geom_line(aes(y = upSum/upNum, colour = "Upstream")) +
  geom_line(aes(y = downSum/downNum, colour = "Downstream")) +
  geom_line(aes(y = exonSum/exonNum, colour = "Exon")) +
#  geom_line(aes(y = intronSum/intronNum, colour = "Intron")) +
  pubTheme + theme(legend.position=c(0.87,0.5)) +
#  ylim(0, 30) +
  labs(x = "NCfem Position (1Mb windows)", 
       y = "Windowed count divided by feature count",
       colour = "Location") +
  facet_grid( vars(scaff), scale = "free_x", space = "free" )

femCounts %>%
  select(c(scaff, winnum, intronNum, intronSum, intronAvg)) %>%
  ggplot(aes(x = winnum)) +
  geom_line(aes(y = intronNum, colour = "Introns")) +
  geom_line(aes(y = intronSum, colour = "TEs in Introns")) +
  pubTheme + theme(legend.position=c(0.87,0.5)) +
  ylim(0, 500) +
  labs(x = "NCfem Position (1Mb windows)", y = "Average in Window",
       colour = "Feature Type") +
  facet_grid( vars(scaff), scale = "free_x", space = "free" )
```



